### 2012-04-18 project-writings/proposal.tex
Add the proposal to this repo.

\documentclass{article}
%\usepackage{times}
\usepackage{url}
\usepackage{amsmath,amsthm,amssymb,color}
%\usepackage{multicol}
\usepackage{float}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
%\usepackage[all]{hypcap}

\begin{document}

\title{\bf Range Analysis for IonMonkey}

\author{
Ryan Pearl\\
\texttt{$<$rpearl@andrew.cmu.edu$>$}\\
\and
Michael Sullivan\\
\texttt{$<$mjsulliv@andrew.cmu.edu$>$}\\
}

\maketitle

\section{Introduction}
We propose implementing and evaluating range analysis for IonMonkey,
an optimizing JIT compiler for Javascript. Range analysis is
particularly useful in this context: in addition to taking advantage
of the analysis results to eliminate redundant comparisons and
bounds-checks, Javascript's dynamic semantics requires that all
numeric computations are floating point arithmetic.  This is slow, so
numbers that can be stored as integers are, and arithmetic operations
are guarded against overflow into their floating point
representation. With range analysis, we can prove these guards
unnecessary and emit higher-quality code.
\section{Project Tracking}
As IonMonkey is an open-source project built by the Mozilla
Corporation, we will use the range analysis tracking bug as our
project page. This will allow discussion from other contributors to
the Javascript engine to be integrated into our project page. The
tracking bug for range analysis is:
\\ \url{https://bugzilla.mozilla.org/show_bug.cgi?id=699883}.

\section{Project Goals}
Our goal is to provide an implementation of range analysis and use the
results to implement two of the potential optimizations we discussed
(eliminating bounds checks, redundant comparisons, and overflow
guards). If progress appears to be too slow, we still hope to complete
range analysis integrate it into the compiler, and maybe implement one
of the optimizations. If our project proceeds faster than expected, we
can implement more optimizations that take advantage of this analysis
or begin to implement other related analysis phases for IonMonkey.

\section{Logistics}
\subsection{Schedule}
\begin{itemize}
\item \textbf{Weeks 1 - 2}:\\
\textit{Research}.
We will go over the various algorithms available for range analysis
and weigh their advantages. In particular, we must take care to
consider the time complexity of our selected algorithm since JIT
compilation must run more quickly than ahead-of-time compilation. At
this point we will be able to see if we can gain any advantage by
using the runtime knowledge available to a JIT compiler.
\item \textbf{Week 2}:\\
\textit{Source-diving}.
IonMonkey is a large code-base and must interact with Firefox's
Javascript runtime and interpreter. While these components are
well-isolated, some source-diving will be necessary to understand the
context of the JIT.

\item \textbf{Weeks 2 - 4}:\\
\textit{Implement analysis}.
Implementing a range analysis phase on the IonMonkey IL and running it
on a body of Javascript code to get an idea of what optimizations are
likely to be most effectively exposed by the information we are
gathering.

\item \textbf{Weeks 4 - 5}:\\
\textit{Implement optimizations}.

Implementing some optimizations that take advantage of range information (such
as eliminating bounds checks, redundant comparisons, and overflow guards)
and testing against the large Javascript test-suite and fuzzer.

\item \textbf{Weeks 5 - 6}:\\
\textit{Evaluation and wrap-up}.

There are several large Javascript benchmarking suites available to
measure the impact of our optimization on the x86, x86\_64, and ARM
architectures. Since one of the other groups is apparently working on
building a new Javascript benchmark suite, we could maybe also try
their benchmark suite.

\end{itemize}
\subsection{Milestones}
By April 19th, roughly the halfway-point of the project, we hope to be
code complete on the analysis phase of the optimization. At this
point, we will begin working on using these analysis results to
optimize code.

\subsection{Literature Search}
We have looked at a few papers on range analysis, such as ``The Design
and Implementation of a Non-Iterative Range Analysis Algorithm on a
Production Compiler'', by Teixeira and Pereira, which is a practical
implementation of ``A class of polynomially solvable range
constraints for interval analysis without widenings.'' by Su and Wagner. The
algorithm due to Su and Wagner is non-iterative, and works by solving a
constraint graph. Other range analyses, generally formulated as a more
traditional iterative dataflow problem, are also available.

\subsection{Resources Needed}
The software we need is just the IonMonkey code base and a handful of
freely available benchmark suites. We both have good, modern computers
to run tests on.

\subsection{Getting Started}
So far, we have done some looking over papers and considerations of
what algorithms to use. Both of us have worked on the Javascript team
at Mozilla, and one of us has worked on IonMonkey, so we should have a
leg up in diving into the engine source. We don't have anything preventing us
from starting immediately.

\end{document}

### 2012-04-18 project-writings/milestone.tex
Write a milestone document.

\documentclass{article}
%\usepackage{times}
\usepackage{url}
\usepackage{amsmath,amsthm,amssymb,color}
\usepackage{fullpage}
%\usepackage{multicol}
\usepackage{float}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
%\usepackage[all]{hypcap}

\begin{document}

\title{\vspace{-1in} \bf Range Analysis for IonMonkey: Milestone}
%\title{\bf Range Analysis for IonMonkey: Milestone}

\author{
Ryan Pearl\\
\texttt{$<$rpearl@andrew.cmu.edu$>$}\\
\and
Michael Sullivan\\
\texttt{$<$mjsulliv@cs.cmu.edu$>$}\\
}

\maketitle

\section{Major Changes}
We have made no major changes to our goals or implementations.

\section{What We Have Accomplished}
We reviewed the literature on different types of range analysis and
decided to implement an analysis based on ``Eliminating Range Checks
using Static Single Assignment Form''. 
One tricky part of this analysis is propagating range information
learned from conditional branches; this requires tracking ranges for
each variable differently at different points in the code, which loses
some of the normal benefits of SSA representations.
This method uses a version of
SSA form augmented with special $\beta$ nodes that act as a copy but
associate range information with the result. These $\beta$ nodes are
inserted after conditional branches to split the variables compared in
the branch and thus propagate the range information learned inside the
branches.

We have implemented passes to augment the middle intermediate
representation (MIR) with $\beta$ nodes and also to remove $\beta$
nodes from the MIR. This passes should be fairly inexpensive, meaning
that we should be able to add $\beta$ nodes, run our optimizations,
and remove $\beta$ nodes at any stage of the optimization pipeline
without requiring changes to other passes.

We have also implemented an initial version of a worklist based SSA
algorithm. The code works, but can not track ranges bounded by other
variables (which is a feature that we may add eventually) and does not yet
support all arithmetic operations.

\section{Meeting Our Milestone}

We think we mostly hit our milestone. We don't have all of the
arithmetic operations implemented, and we had hoped to but that is
just legwork that should fit easily into our existing framework.
The reason we are a little behind is because figuring out what
algorithms we wanted to implement and what direction to take took
longer than expected. We spent a lot of time trying to figure out ways
we could avoid having to augment the SSA form before we realized that
it wouldn't be that hard and would greatly simplify matters.

\section{Surprises}

We haven't run into any major surprises. We were a little surprised by
how much work has been done on range analysis that is not applicable
to what we want to do.

\section{Logistics}
\subsection{Schedule}
\begin{itemize}
\item \textbf{Week 5}:\\
\textit{Implement optimizations, finish analysis}.

The current plan is for Ryan to work on wrapping up the analysis while
Michael starts implementing optimizations (probably starting with
eliminating overflow guards). When the analysis is wrapped up, Ryan
will join in writing optimizations.

\item \textbf{Week 6}:\\
\textit{Evaluation and wrap-up}.

We will do benchmarks using common Javascript benchmark suites and
tune our optimizations accordingly.

\end{itemize}

\subsection{Resources Needed}
Nothing.

\end{document}

### 2012-04-19 sync.sh
exclude project-writings from sync

@@ -1,4 +1,4 @@
 #!/bin/sh
 
 jekyll
-rsync -vaz --exclude='*.sh' _site/ cmplrz@www.endofunctor.org:~/public_html/
+rsync -vaz --exclude='*.sh' --exclude='project-writings/*' _site/ cmplrz@www.endofunctor.org:~/public_html/

### 2012-04-29 project-writings/paper.tex
Produce a skeleton of the report.

--- 2012-04-18 project-writings/proposal.tex
@@
 \documentclass{article}
+\usepackage{fullpage}
 %\usepackage{times}
@@
 
-\title{\bf Range Analysis for IonMonkey}
+\title{\bf Range Analysis for the IonMonkey JavaScript engine}
 
@@
 Michael Sullivan\\
-\texttt{$<$mjsulliv@andrew.cmu.edu$>$}\\
+\texttt{$<$mjsulliv@cs.cmu.edu$>$}\\
 }
@@
 \section{Introduction}
+%% XXX: FIXME, TODO
 We propose implementing and evaluating range analysis for IonMonkey,
@@
 unnecessary and emit higher-quality code.
-\section{Project Tracking}
-As IonMonkey is an open-source project built by the Mozilla
-Corporation, we will use the range analysis tracking bug as our
-project page. This will allow discussion from other contributors to
-the Javascript engine to be integrated into our project page. The
-tracking bug for range analysis is:
-\\ \url{https://bugzilla.mozilla.org/show_bug.cgi?id=699883}.
 
-\section{Project Goals}
-Our goal is to provide an implementation of range analysis and use the
-results to implement two of the potential optimizations we discussed
-(eliminating bounds checks, redundant comparisons, and overflow
-guards). If progress appears to be too slow, we still hope to complete
-range analysis integrate it into the compiler, and maybe implement one
-of the optimizations. If our project proceeds faster than expected, we
-can implement more optimizations that take advantage of this analysis
-or begin to implement other related analysis phases for IonMonkey.
-
-\section{Logistics}
-\subsection{Schedule}
-\begin{itemize}
-\item \textbf{Weeks 1 - 2}:\\
-\textit{Research}.
-We will go over the various algorithms available for range analysis
-and weigh their advantages. In particular, we must take care to
-consider the time complexity of our selected algorithm since JIT
-compilation must run more quickly than ahead-of-time compilation. At
-this point we will be able to see if we can gain any advantage by
-using the runtime knowledge available to a JIT compiler.
-\item \textbf{Week 2}:\\
-\textit{Source-diving}.
-IonMonkey is a large code-base and must interact with Firefox's
-Javascript runtime and interpreter. While these components are
-well-isolated, some source-diving will be necessary to understand the
-context of the JIT.
-
-\item \textbf{Weeks 2 - 4}:\\
-\textit{Implement analysis}.
-Implementing a range analysis phase on the IonMonkey IL and running it
-on a body of Javascript code to get an idea of what optimizations are
-likely to be most effectively exposed by the information we are
-gathering.
-
-\item \textbf{Weeks 4 - 5}:\\
-\textit{Implement optimizations}.
-
-Implementing some optimizations that take advantage of range information (such
-as eliminating bounds checks, redundant comparisons, and overflow guards)
-and testing against the large Javascript test-suite and fuzzer.
-
-\item \textbf{Weeks 5 - 6}:\\
-\textit{Evaluation and wrap-up}.
-
-There are several large Javascript benchmarking suites available to
-measure the impact of our optimization on the x86, x86\_64, and ARM
-architectures. Since one of the other groups is apparently working on
-building a new Javascript benchmark suite, we could maybe also try
-their benchmark suite.
-
-\end{itemize}
-\subsection{Milestones}
-By April 19th, roughly the halfway-point of the project, we hope to be
-code complete on the analysis phase of the optimization. At this
-point, we will begin working on using these analysis results to
-optimize code.
-
-\subsection{Literature Search}
+\section{Related Work}
+%% XXX: FIXME, TODO
 We have looked at a few papers on range analysis, such as ``The Design
@@
 Production Compiler'', by Teixeira and Pereira, which is a practical
-implementation of ``A class of polynomially solvable range
-constraints for interval analysis without widenings.'' by Su and Wagner. The
-algorithm due to Su and Wagner is non-iterative, and works by solving a
-constraint graph. Other range analyses, generally formulated as a more
-traditional iterative dataflow problem, are also available.
+implementation of an algorithm designed by by Su and Wagner
+\cite{Su04aclass}. The algorithm due to Su and Wagner is
+non-iterative, and works by solving a constraint graph. Other range
+analyses, generally formulated as a more traditional iterative
+dataflow problem, are also available.
 
-\subsection{Resources Needed}
-The software we need is just the IonMonkey code base and a handful of
-freely available benchmark suites. We both have good, modern computers
-to run tests on.
+We reviewed the literature on different types of range analysis and
+decided to implement an analysis based on Gough and Klaeren's SSA
+based range analysis algorithm \cite{Gough94eliminatingrange}.
 
-\subsection{Getting Started}
-So far, we have done some looking over papers and considerations of
-what algorithms to use. Both of us have worked on the Javascript team
-at Mozilla, and one of us has worked on IonMonkey, so we should have a
-leg up in diving into the engine source. We don't have anything preventing us
-from starting immediately.
+\section{Implementation}
+%% XXX: TODO
+
+\section{Evaluation}
+%% XXX: TODO
+
+\section{Surprises and Lessons Learned}
+%% XXX: TODO
+
+\section{Conclusions}
+%% XXX: TODO
+Boy, JS is a pain in the ass.
+
+\bibliography{citations}{}
+\bibliographystyle{abbrv}
 

### 2012-04-29 project-writings/paper.tex
Write some intro stuff.

@@
 \section{Introduction}
-%% XXX: FIXME, TODO
-We propose implementing and evaluating range analysis for IonMonkey,
-an optimizing JIT compiler for Javascript. Range analysis is
-particularly useful in this context: in addition to taking advantage
-of the analysis results to eliminate redundant comparisons and
-bounds-checks, Javascript's dynamic semantics requires that all
-numeric computations are floating point arithmetic.  This is slow, so
-numbers that can be stored as integers are, and arithmetic operations
-are guarded against overflow into their floating point
-representation. With range analysis, we can prove these guards
-unnecessary and emit higher-quality code.
+With the explosion of browser based applications that has occured in
+recent years, Javascript has become one of the most widely used
+programming languages, and Javascript performance has become
+critically important. Unfortunately, Javascript seems to be designed
+to be un-performant.
+
+One source of these woes are Javascript's dynamic semantics for
+numbers: all numeric values are defined by the specification to be
+IEEE 754 double precision floating point values (with bitwise
+operations being defined to round to the nearest 32 bit integer before
+performing the operation). Since the common case is that numeric
+values are integers and because floating point operations are
+more expensive, all major Javascript implementations attempt to
+represent numbers as integers when possible.
+
+In IonMonkey, Mozilla's next-generation optimizing Javascript JIT
+compiler, an approximate type inference system is used to generate
+type-specialized code for the common case of integer arithmetic, and
+type guards are inserted if the type inferencer is not certain. The
+catch is that since double precision floats have a larger range than
+32-bit integers, if the integer operations would overflow the result
+must be promoted to being stored in a double. IonMonkey deals with
+this by inserting overflow checks after arithmetic operations and
+bailing out of the type specialized code if they trip. While this is
+sometimes actually necessary, it is frequently the case that overflow
+is not possible.
+
+We have implemented integer value range analysis in the IonMonkey
+engine and used it to eliminate provably unnecessary overflow checks.
 

### 2012-04-29 project-writings/paper.tex
More things

@@
 engine and used it to eliminate provably unnecessary overflow checks.
+We implemented an SSA-based worklist range analysis algorithm that
+propagates integer ranges for variables by performing range
+arithmetic. The algorithm is based off of Gough and Klaeren's range
+analysis for bounds check elimination
+\cite{Gough94eliminatingrange}. A range (including information about
+possible overflow) is associated with each computation, and the ranges
+are consulted to determine whether to perform overflow checks for
+arithmetic computations.
+
+OTHER OPTIMIZATIONS WE DO
+
+OUR CONTRIBUTIONS
 

### 2012-04-30 project-writings/paper.tex
Write up some of the equations for the stuff we do.
Write a lot of paper

@@
 %\usepackage[all]{hypcap}
+
+\newcommand{\Squash}{\mathsf{Squash}}
+\newcommand{\Clamp}{\mathsf{Clamp}}
+\newcommand{\Range}{\mathsf{Range}}
+\let\x\cdot
 
@@
 
-OTHER OPTIMIZATIONS WE DO
+OTHER OPTIMIZATIONS WE DO??
 
-OUR CONTRIBUTIONS
+The core of our analysis system is essentially an implementation of
+Gough and Klaeren's static range analysis algorithm for bounds check
+elimination.
+We identify our technical contributions as:
+\begin{itemize}
+\item An implementation of Gough and Klaeren's range analysis in a
+  modern industrial-strength SSA-based compiler.
+\item A different algorithm for inserting and removing $\beta$ nodes
+  that used by Gough and Klaeren that takes advantage of the data
+  structures of a modern SSA compiler.
+\item To facilitate the above, extended IonMonkey with a fast
+  (constant time, one integer per basic block memory overhead) 
+  mechanism for determining whether a block is dominated by another.
+\item Extending the analysis to produce the information needed for
+  overflow check elimination by differentiating between
+  ``maximum/minimum 32-bit int values'' and ``infinity'' (possible
+  overflow).
+\item The observation that when doing range computation with the above
+  distinction, the input ranges can be always be ``squashed'' to only
+  finite ranges, since if overflow had occured, the code would have
+  bailed out.
+\item An implementation of overflow check elimination using the above
+  analyses.
+\end{itemize}
 
@@
 
+\section{IonMonkey Overview}
+
 \section{Implementation}
 %% XXX: TODO
+
+\subsection{Ranges and Range Arithmetic}
+
+In Single Static Assignment form, each computation has a distinct
+name. We take advantage of this to assign each name (that is, each
+computation) a ``range''; the ranges in our system are a contiguous
+subset of the set ${-\infty} \cup [-2^{31}, 2^{31}-1] \cup {\infty}$
+that represent the possible results a computation may have. A range
+that includes $\pm \infty$ indicates that the associated
+computation may overflow in that direction.
+
+\begin{figure}[ht]
+\begin{eqnarray*}
+Z &=& {-\infty} \cup [-2^{31}, 2^{31}-1] \cup {\infty} \\
+\Clamp(n) &=& \begin{cases}
+-\infty &\text{if } n < -2^{31} \\
+\infty &\text{if } n > -2^{31}-1 \\
+n&\text{otherwise } \\
+\end{cases}\\
+\Clamp([l, h]) &=& [\Clamp(l), \Clamp(h)] \\
+{[x_l, x_h] \cup [y_l, y_h]} &=& [\min(x_l, y_l), \max(x_h, y_h)] \\
+{[x_l, x_h] \cap [y_l, y_h]} &=& [\max(x_l, y_l), \min(x_h, y_h)] \\
+{[x_l, x_h] + [y_l, y_h]} &=& \Clamp ([x_l + y_l, x_h + y_h]) \\
+{[x_l, x_h] - [y_l, y_h]} &=& \Clamp ([x_l - y_h, x_h + y_l]) \\
+{[x_l, x_h] \x [y_l, y_h]} &=& 
+\begin{tabular}{rl}
+$\Clamp ([$&$\min(x_l \x y_l, x_l \x y_h, x_h \x y_l, x_h \x y_h),$\\
+&$\max(x_l \x y_l, x_l \x y_h, x_h \x y_l, x_h \x y_h)])$ \\
+\end{tabular}
+\end{eqnarray*}
+
+\caption{Range arithmetic definitions}
+\label{fig:range_arith}
+\end{figure}
+
+We can define arithmetical and set operations on these ranges in the
+standard way, ``clamping'' values outside the range of a 32-bit signed
+integer to $\pm \infty$. Some definitions for this are are shown in
+Figure \ref{fig:range_arith}.
+
+\subsection{Conditional Branches and $\beta$ nobes}
+
+An important source of range information that requires care to take
+advantage of is conditional branches. Consider the code in Figure
+\ref{fig:branch}. All three arithmetic operations in this code can be
+proven to not overflow, but to do this it is not sufficient to simply
+associate each name with a range, since the information differs
+between basic blocks. An obvious approach is to instead associate
+ranges with name/basic block pairs, allowing representation of
+flow-sensitive range information. This solution, however, is
+unsatisfying. One of the major benefits of SSA-style analysis is
+eliminating the need for this sort of tracking; by giving each
+computation a unique name, SSA reduces the need for tracking
+information about variables across time; adopting this approach would
+undo this.
+
+The approach we took, again following \cite{Gough94eliminatingrange},
+is to instead attack the problem in ``SSA style'', by introducing
+$\beta$-nodes, a new form of pseudo-operation that associates range
+information with a value. $\beta$ nodes take one argument and
+additionall have an auxillary constant range associated with
+them. Operationally, $\beta$-nodes are simply copies, but the
+invariant is maintained that the output value will fall inside the
+range given by the $\beta$ node. Gough and Klaeren refer to SSA
+extended with $\beta$ nodes as XSA.
+
+\begin{figure}[ht]
+\begin{center}
+\begin{verbatim}
+var y;
+if (x < 0) {
+    y = x + 2000000000;
+} else {
+    if (x < 1000000000) {
+        y = x*2;
+    } else {
+        y = x - 3000000000;
+    }
+}
+\end{verbatim}
+\end{center}
+\caption{Code that requires information about branches to properly analyze}
+\label{fig:branch}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\begin{verbatim}
+if (x < 0) {
+    x1 = Beta(x, [INT_MIN, -1]);
+    y1 = x1 + 2000000000;
+} else {
+    x2 = Beta(x, [0, INT_MAX]);
+    if (x2 < 1000000000) {
+        x3 = Beta(x2, [INT_MIN, 999999999]);
+        y2 = x3*2;
+    } else {
+        x4 = Beta(x2, [1000000000, INT_MAX]);
+        y3 = x4 - 3000000000;
+    }
+    y4 = Phi(y2, y3);
+}
+y = Phi(y1, y4);
+\end{verbatim}
+\end{center}
+\caption{Code from Figure \ref{fig:branch}, in XSA form}
+\label{fig:xsa}
+\end{figure}
+
+\subsection{Range Analysis}
+
+\begin{figure}[ht]
+\begin{eqnarray*}
+\Squash(n) &=& \begin{cases}
+ -2^{31} &\text{if } n = -\infty \\
+2^{31}-1 &\text{if } n = \infty \\
+n&\text{otherwise } \\
+\end{cases}\\
+\Squash([l, h]) &=& [\Squash(l), \Squash(h)]
+\end{eqnarray*}
+%
+\begin{eqnarray*}
+x = \Phi(y, z) &\Rightarrow&
+    \Range(x) = \Squash(\Range(y)) \cup \Squash(\Range(z)) \\
+x = \beta(y, [l, h]) &\Rightarrow& \Range(x) = \Squash(\Range(y)) \cap [l, h] \\
+x = y + z &\Rightarrow& \Range(x) = \Squash(\Range(y)) + \Squash(\Range(z)) \\
+x = y - z &\Rightarrow& \Range(x) = \Squash(\Range(y)) - \Squash(\Range(z)) \\
+x = y \x z &\Rightarrow& \Range(x) = \Squash(\Range(y)) \x \Squash(\Range(z)) \\
+x = c &\Rightarrow& \Range(x) = [c, c] \\
+\end{eqnarray*}
+
+\caption{Range analysis equations}
+\label{fig:range_anal}
+\end{figure}
 

### 2012-04-30 project-writings/paper.tex
Spell check.

@@
 \section{Introduction}
-With the explosion of browser based applications that has occured in
+With the explosion of browser based applications that has occurred in
 recent years, Javascript has become one of the most widely used
@@
 numbers: all numeric values are defined by the specification to be
-IEEE 754 double precision floating point values (with bitwise
+IEEE 754 double precision floating point values (with bit-wise
 operations being defined to round to the nearest 32 bit integer before
@@
   distinction, the input ranges can be always be ``squashed'' to only
-  finite ranges, since if overflow had occured, the code would have
+  finite ranges, since if overflow had occurred, the code would have
   bailed out.
@@
 information with a value. $\beta$ nodes take one argument and
-additionall have an auxillary constant range associated with
+additionally have an auxiliary constant range associated with
 them. Operationally, $\beta$-nodes are simply copies, but the

### 2012-04-30 project-writings/paper.tex
More paper. 5th page!

@@
 
+\subsection{Inserting and Removing Beta Nodes}
+To put programs in XSA form, Gough and Klaeren presented an extension
+to Cytron et al.'s standard algorithm for computing SSA form using
+dominance frontiers \cite{Cytron:1991:ECS:115372.115320}.
+
+This approach does not work for us for two reasons: first, IonMonkey
+does not use Cytron et al.'s algorithm for computing SSA form; it
+instead takes advantage of the structured nature of Javascript code to
+go to SSA form without computing dominance frontiers \footnote{More
+  precisely, it takes advantage in an ad-hoc and terrifying way of
+  poorly documented structured invariants and ``source notes'' in the
+  bytecode that the SpiderMonkey frontend emits.}.  Furthermore, we
+decided that instead of putting the program into XSA form at the
+beginning and leaving it there through the compilation phase, it would
+be a better idea to introduce $\beta$ nodes solely for performing
+range analysis and remove them afterwards. This is for two main
+reasons: first, as a question of engineering effort, we did not want
+to teach every pass of the compiler how to deal with $\beta$ nodes,
+and the IonMonkey team does not want us to change the IL in a pervasive
+way simply for range analysis; second, $\beta$ nodes are copies that
+are not allowed to be propagated away, which may inhibit optimization
+passes (like global value numbering).
+
+Thus we developed a quite simple and in practice efficient algorithm to
+add $\beta$ nodes to an SSA source program by taking
+advantage of data structures common to modern SSA based compilers.
+\footnote{The method for removing $\beta$ nodes is so trivial that we
+  could not claim with a straight face that we ``developed an
+  algorithm'' to perform it.}
+
 \subsection{Range Analysis}
@@
 x = y \x z &\Rightarrow& \Range(x) = \Squash(\Range(y)) \x \Squash(\Range(z)) \\
-x = c &\Rightarrow& \Range(x) = [c, c] \\
+x = c &\Rightarrow& \Range(x) = [c, c]
 \end{eqnarray*}
-
 \caption{Range analysis equations}

### 2012-04-30 project-writings/paper.tex
Add pseudocode for the beta node insertion algorithm.

@@
 
+\begin{figure}[ht]
+\begin{center}
+\begin{verbatim}
+function addBetaNodesToFunc(cfg):
+  for each b in cfg.basic-blocks():
+    if b.numPredecessors() == 1 and isBranch(b.pred(0).lastInstruction())
+      test = b.pred(0).lastInstruction()
+      # find whether we are the true or false branch
+      direction = (b == test.trueTarget())
+      if test is of the form <x RELOP c>:
+          r = computeRange(RELOP, c, direction)
+          x' = b.insertFront(<Beta(x, r)>)
+          replaceDominatedUses(x, x')
+
+# Replace all uses of the instruction x that are dominated by
+# the instruction x' with x'
+function replaceDominatedUses(x, x'):
+  for each use in x.uses():
+    ins = use.instruction()
+    if ins != x' and isDominatedUse(x', use):
+      replaceUseWith(use, x')
+
+function isDominatedUse(x', use):
+  ins = use.instruction()
+  if ins.isPhiNode():
+    # Get the predecessor block that corresponds to the
+    # argument of the phi node we are looking at
+    phi-arg-pred-block = ins.block().pred(use.index())
+    return blockDominates(x'.block(), phi-arg-pred-block)
+  else:
+    return blockDominates(x'.block(), x.block())
+\end{verbatim}
+\end{center}
+\caption{Algorithm for inserting beta nodes}
+\label{fig:add_betas}
+\end{figure}
+
 \subsection{Range Analysis}

### 2012-04-30 project-writings/paper.tex
Lots more paper work.

@@
 add $\beta$ nodes to an SSA source program by taking
-advantage of data structures common to modern SSA based compilers.
+advantage of data structures common to modern SSA based compilers
 \footnote{The method for removing $\beta$ nodes is so trivial that we
   could not claim with a straight face that we ``developed an
-  algorithm'' to perform it.}
+  algorithm'' to perform it.} .  Pseudocode for the algorithm is shown
+in Figure \ref{fig:add_betas}. The algorithm traverses the control
+flow group, looking for blocks that have been entered by a jump
+conditioned on a relational operation. When we find a variable being
+compared to a constant, we compute a range and use it to create a new
+$\beta$ node. We want to propagate this range information to all uses
+of the variable where the branch is guarenteed to have been taken;
+this is accomplished by replacing all uses of original instruction
+that are dominated by the new $\beta$ instruction with the $\beta$. To
+do this, we take advantage of IonMonkey's use chains to efficiently
+iterate over all uses of an instruction.
+
+In order to do this, we needed to be able to efficiently determine
+whether one block is dominated by another. The best way to do that
+that existed in IonMonkey was to walk up the dominator tree looking
+for the block being tested against. We felt that it would be too
+inefficient to need to do this for every use of every variable tested
+with a conditional. To do this, we annotated each basic block with its
+index in a pre-order traversal of the dominator tree. Since IonMonkey
+basic blocks already tracked how many other blocks they dominated, we
+can now easily check whether a block dominates another by checking
+whether the potential descendent's index falls within the range of the
+ancestor's dominated blocks.
 
@@
     return blockDominates(x'.block(), x.block())
+
+function blockDominates(b1, b2):
+  high = b1.domIndex() + b1.numDominated()
+  low  = b1.domIndex()
+  return b2.domIndex() >= low && b2.domIndex <= high
 \end{verbatim}

### 2012-04-30 project-writings/paper.tex
Make a note about removing nobes
spellcheck

@@
   poorly documented structured invariants and ``source notes'' in the
-  bytecode that the SpiderMonkey frontend emits.}.  Furthermore, we
+  bytecode that the SpiderMonkey front-end emits.}.  Furthermore, we
 decided that instead of putting the program into XSA form at the
@@
 $\beta$ node. We want to propagate this range information to all uses
-of the variable where the branch is guarenteed to have been taken;
+of the variable where the branch is guaranteed to have been taken;
 this is accomplished by replacing all uses of original instruction
@@
 can now easily check whether a block dominates another by checking
-whether the potential descendent's index falls within the range of the
+whether the potential descendant's index falls within the range of the
 ancestor's dominated blocks.
@@
 
+The method for removing beta nodes is trivial. Traverse the program
+removing beta nodes and replacing all uses of them with their
+arguments. Since beta nodes only are placed at the beginning of basic
+blocks, we only need to iterate through the $\beta$ node prefix of a
+block.
+
 \subsection{Range Analysis}

### 2012-04-30 project-writings/paper.tex
Cite type inference paper

@@
 In IonMonkey, Mozilla's next-generation optimizing Javascript JIT
-compiler, an approximate type inference system is used to generate
+compiler, a hybrid type inference system (which uses both static analysis and
+runtime information\cite{BHackettTI}) is used to generate
 type-specialized code for the common case of integer arithmetic, and

### 2012-04-30 project-writings/paper.tex
Update related work a bit

@@
 %% XXX: FIXME, TODO
-We have looked at a few papers on range analysis, such as ``The Design
+There are a variety of approaches to range analysis. A recent work, ``The Design
 and Implementation of a Non-Iterative Range Analysis Algorithm on a
-Production Compiler'', by Teixeira and Pereira, which is a practical
+Production Compiler'', by Teixeira and Pereira, outlines a practical implementation of
 implementation of an algorithm designed by by Su and Wagner
 \cite{Su04aclass}. The algorithm due to Su and Wagner is
-non-iterative, and works by solving a constraint graph. Other range
-analyses, generally formulated as a more traditional iterative
-dataflow problem, are also available.
+non-iterative, and works by constructing a constraint graph from the program's control flow graph.
+
+A variety of other range analyses are available which formulate the problem as
+a more traditional iterative dataflow problem. GCC, for example, uses an SSA
+workflow algorithm due to Patterson in ``Accurate Static Branch Prediction by
+Value Range Propagation.'' The orignal non-SSA based dataflow algorithm is by
+Harrison in ``Compiler Analysis of the Value Ranges of Variables.''
 

### 2012-04-30 project-writings/paper.tex
Talk about ionmonkey

@@
 %% XXX: TODO
+The IonMonkey compiler is somewhat loosely based on LLVM, in that it is an
+imperative SSA compiler. IonMonkey takes the bytecode used by the SpiderMonkey
+Javascript interpreter and compiles it to native code. Optimistic type
+specialization is used, and if type assumptions fail, then execution ``bails
+out'' to the interpreter.
+
+IonMonkey has two intermediate languages. There is a ``Middle Intermediate
+Representation'' which encompasses higher-level constructs such as lambdas,
+objects, and strings. A variety of optimizations are performed, and then the
+code is lowered to a ``Lower Intermediate Representation'' used for register
+allocation and code generation.
+
+While the ECMA specification for Javascript requires that all numbers behave as
+double-precision floating point values, for performance reasons JIT compilers
+represent numbers as 32-bit integers when possible. However, since doubles can
+represent larger values than integers, operations must be guarded against
+overflow.  Our range analysis pass eliminates overflow checks when operations
+can be proven not to overflow.
 

### 2012-05-01 project-writings/paper.tex
our pseudocode is valid, syntax highlighted python

@@
 \usepackage{hyperref}
+\usepackage{courier}
 \usepackage{listings}
@@
 \newcommand{\Range}{\mathsf{Range}}
+
 \let\x\cdot
+\lstset{
+  language=Python,
+  showstringspaces=false,
+  tabsize=4,
+  basicstyle=\ttfamily,
+  keywordstyle=\bfseries\ttfamily,
+  captionpos=b
+}
 
@@
   algorithm'' to perform it.} .  Pseudocode for the algorithm is shown
-in Figure \ref{fig:add_betas}. The algorithm traverses the control
+in Listing \ref{lst:add_betas}. The algorithm traverses the control
 flow group, looking for blocks that have been entered by a jump
@@
 
-\begin{figure}[ht]
-\begin{center}
-\begin{verbatim}
-function addBetaNodesToFunc(cfg):
-  for each b in cfg.basic-blocks():
-    if b.numPredecessors() == 1 and isBranch(b.pred(0).lastInstruction())
+\begin{lstlisting}[language=Python,
+                   caption={Pseudocode algorithm for inserting beta nodes},
+                   label={lst:add_betas}]
+def addBetaNodesToFunc(cfg):
+  for b in cfg.basic_blocks():
+    if b.numPredecessors() == 1 and isBranch(b.pred(0).lastInstruction()):
       test = b.pred(0).lastInstruction()
@@
       direction = (b == test.trueTarget())
-      if test is of the form <x RELOP c>:
-          r = computeRange(RELOP, c, direction)
-          x' = b.insertFront(<Beta(x, r)>)
-          replaceDominatedUses(x, x')
+      if test.op() in RELOPs:
+          r = computeRange(test.op(), c, direction)
+          x_prime = b.insertFront(Beta(x, r))
+          replaceDominatedUses(x, x_prime)
 
 # Replace all uses of the instruction x that are dominated by
-# the instruction x' with x'
-function replaceDominatedUses(x, x'):
+# the instruction x_prime with x_prime
+def replaceDominatedUses(x, x_prime):
   for each use in x.uses():
     ins = use.instruction()
-    if ins != x' and isDominatedUse(x', use):
-      replaceUseWith(use, x')
+    if ins != x_prime and isDominatedUse(x_prime, use):
+      replaceUseWith(use, x_prime)
 
-function isDominatedUse(x', use):
+def isDominatedUse(x_prime, use):
   ins = use.instruction()
@@
     # argument of the phi node we are looking at
-    phi-arg-pred-block = ins.block().pred(use.index())
-    return blockDominates(x'.block(), phi-arg-pred-block)
+    phi_arg_pred_block = ins.block().pred(use.index())
+    return blockDominates(x_prime.block(), phi_arg_pred_block)
   else:
-    return blockDominates(x'.block(), x.block())
+    return blockDominates(x_prime.block(), x.block())
 
-function blockDominates(b1, b2):
+def blockDominates(b1, b2):
   high = b1.domIndex() + b1.numDominated()
   low  = b1.domIndex()
-  return b2.domIndex() >= low && b2.domIndex <= high
-\end{verbatim}
-\end{center}
-\caption{Algorithm for inserting beta nodes}
-\label{fig:add_betas}
-\end{figure}
+  return b2.domIndex() >= low and b2.domIndex <= high
+\end{lstlisting}
 

### 2012-05-01 project-writings/paper.tex
if we're going to use valid python, it should be valid python. I'm done now.

@@
 def replaceDominatedUses(x, x_prime):
-  for each use in x.uses():
+  for use in x.uses():
     ins = use.instruction()
@@
   low  = b1.domIndex()
-  return b2.domIndex() >= low and b2.domIndex <= high
+  return b2.domIndex() >= low and b2.domIndex() <= high
 \end{lstlisting}

### 2012-05-01 project-writings/paper.tex
add a note for rpearl
Some notes about things

@@
 that includes $\pm \infty$ indicates that the associated
-computation may overflow in that direction.
+computation may overflow in that direction. This distinction between
+the minimum/maximum 32-bit integer values and $\pm \infty$ is the key
+to detecting possible overflow.
 
@@
 
+At the start of range analysis, all ranges are initialized to
+$[-\infty, \infty]$.
+
+XXX: rpearl, explain how the analysis pass works; specifically, when
+we add things to the worklist, because that is different from the
+paper. maybe add pseudocode
+
+Recomputation of the possible range of an instruction is fairly simple
+and proceeds by the rules given in Figure \ref{fig:range_anal}. The
+key insight we had in these equations is that while the ranges of the
+inputs to an operation may be infinite (that is, they may have
+overflowed), this possibility can be ignored when using the ranges as
+inputs. This is because had the computation overflowed, it would have
+been detected and IonMonkey would bail out to non-integer code. We
+know that if code that assumes the input is an integer is still being
+run, then the operation must not have overflowed, and thus must have a
+finite range. What this means for the range analysis equations is that
+we can always ``squash'' the ranges of inputs by replacing positive
+and negative infinite with the maximum and minimum 32-bit integers,
+respectively.
+
 \begin{figure}[ht]

### 2012-05-01 project-writings/paper.tex
Move some text to IonMonkey overview
Talk about the optimizations

@@
 \section{IonMonkey Overview}
-
-\section{Implementation}
-%% XXX: TODO
 The IonMonkey compiler is somewhat loosely based on LLVM, in that it is an
@@
 
-IonMonkey has two intermediate languages. There is a ``Middle Intermediate
-Representation'' which encompasses higher-level constructs such as lambdas,
-objects, and strings. A variety of optimizations are performed, and then the
-code is lowered to a ``Lower Intermediate Representation'' used for register
-allocation and code generation.
+IonMonkey has two intermediate languages. There is a ``Middle
+Intermediate Representation'' (MIR) which encompasses higher-level
+constructs such as lambdas, objects, and strings. A variety of
+optimizations are performed, and then the code is lowered to a ``Lower
+Intermediate Representation'' (LIR) used for register allocation and
+code generation.
 
@@
 can be proven not to overflow.
+
+\section{Implementation}
 
@@
 
+\subsection{Overflow Check Elimination}
+Overflow check elimination is implemented as part of the lowering pass
+that generates the lower level intermediate representation (LIR) from
+the middle intermediate representation (MIR). During lowering,
+operations that can fail in such a way as to require a bailout have
+``snapshots'' associated with them. During actual code generation,
+instruction-specific checking code (overflow checking, in this case)
+is emitted for operations that can fail and have snapshots associated
+with them.
+
+During lowering of arithmetic operations that can overflow, we check
+whether the instructions range is finite; if it is, we supress
+assigning a snapshot to the operation, eliminating the overflow check.
+
+\subsection{Array Bounds Check Elimination}
+We do not perform general array bounds check elimination. To do this
+fully would require symbolic range analysis. Since IonMonkey already
+does hoisting of bounds checks out of loops, we determined that the
+effort would probably not be worth the gain.
+
+We do eliminate bounds checks in one situation: when bounds checks
+have been hoisted out of a loop, a check is added that specifically
+bounds checks whether a value is greater than the array minimum index
+(which, unfortunately, need not be zero). During lowering we will
+eliminate this check if we can prove that the index is greater than
+the minimum value.
+
 \section{Evaluation}

### 2012-05-01 project-writings/paper.tex
explain the range propagation algorithm

@@
 
-XXX: rpearl, explain how the analysis pass works; specifically, when
-we add things to the worklist, because that is different from the
-paper. maybe add pseudocode
+Our algorithm for range analsis is based on the algorithm due to Gough. It
+differs from the paper in a few ways to better fit the architecture of
+IonMonkey. The paper suggests using a dominator tree order, but we found that
+using reverse post-order was better for our purposes: reverse post-order
+traversal is simpler to implement in IonMonkey, and in general definitions are
+still visited before all their uses. There is no issue of correctness since we
+iterate until no range information is changing..
+
+We compute the range of each definition unconditionally once, and add all phi
+nodes and beta nodes to a worklist--the only values which can change are those
+that flow through backedges captured by phis. We then iterate through the
+worklist, recomputing the range of each definition, and add all consumers of
+that definition if the range changed in order to propagate this information.
+
+\begin{lstlisting}[language=Python,
+                   caption={Pseudocode algorithm for propagating range information},
+                   label={lst:propagate_ranges}]
+def propagateRanges(graph):
+    worklist = []
+    for block in reversePostOrder(graph):
+        for insn in block:
+            insn.recomputeRange()
+            if !insn.isPhi() and !insn.isBeta():
+                continue
+            worklist.appendAll(insn.uses())
+
+    while len(worklist) > 0:
+        insn = worklist.pop()
+        #recomputeRange() returns true if the range changed
+        if (insn.recomputeRange()):
+            worklist.appendAll(insn.uses())
+\end{lstlisting}
 

### 2012-05-01 project-writings/paper.tex
The pseudocode doesn't need to be fucking python

@@
       direction = (b == test.trueTarget())
-      if test.op() in RELOPs:
-          r = computeRange(test.op(), c, direction)
+      if test matches ``x RELOP c'':
+          r = computeRange(RELOP, c, direction)
           x_prime = b.insertFront(Beta(x, r))

### 2012-05-01 project-writings/paper.tex
Talk about our algorithm more

@@
 still visited before all their uses. There is no issue of correctness since we
-iterate until no range information is changing..
+iterate until no range information is changing.
 
@@
 that definition if the range changed in order to propagate this information.
+
+SSA form lets us use a ``sparse'' analysis in which we only recompute
+information that changes. This furthermore means that the runtime of our
+SSA-based algorithm is bounded by the number of loop backedges in the graph: in
+a graph with no back edges, on the second iteration we will discover that no
+range information has changed and terminate the pass.
+
+This method for performing range propagation has the added benefit that at
+every iteration of the algorithm, the range information is correct. Additional
+iterations serve only to yield more information. This is beneficial to
+IonMonkey since there are plans to implement a ``base-line'' compiler which
+performs small, cheap optimizations such as pessimistic global value numbering,
+which runs in a single pass. When code becomes hot, it can be recompiled with
+more expensive optimziations enabled. We will be able to run range analysis for
+a few iterations in the base-line compilation mode, and run to a fixed point
+when performing more aggressive optimization.
 

### 2012-05-02 project-writings/paper.tex
cite the ECMA spec

@@
 While the ECMA specification for Javascript requires that all numbers behave as
-double-precision floating point values, for performance reasons JIT compilers
+double-precision floating point values\cite{ECMA-262}, for performance reasons JIT compilers
 represent numbers as 32-bit integers when possible. However, since doubles can

### 2012-05-02 project-writings/paper.tex
expound a bit upon beta nobes
rename Cytron citation since vim's make parsing was getting confused when there was an error

@@
 range given by the $\beta$ node. Gough and Klaeren refer to SSA
-extended with $\beta$ nodes as XSA.
+extended with $\beta$ nodes as XSA. These $\beta$ nodes are used to split
+definitions at conditionally taken branches, and augment uses inside the
+branches with range information.
 
@@
 to Cytron et al.'s standard algorithm for computing SSA form using
-dominance frontiers \cite{Cytron:1991:ECS:115372.115320}.
+dominance frontiers \cite{Cytron}.
 

### 2012-05-02 project-writings/paper.tex
benchmark tabular results

@@
 \usepackage{float}
+\usepackage[section]{placeins}
 \usepackage{graphicx}
@@
 \section{Evaluation}
+To evaluate our work, we tested for correctness using Mozilla's large
+repository of regression tests, and measured the impact of our
+optimization by the commonly available benchmark suites.
+%% TODO : elim stats too
+
+There are three benchmark suites which are generally used to measure Javascript
+JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
+suite developed by Apple, and the Kraken benchmark suite developed by Mozilla.
+
+We ran each benchmark suite on IonMonkey, both with and without our range
+analysis pass, and measured the performance differences. The results are
+summarized below:
+
+\begin{figure}[H]
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+    \textbf{Test}     &\textbf{Comparison} & \textbf{From}        & \textbf{To}           &  \textbf{Details} \\
+\hline\hline
+    Total:            &   -                & 3414.7ms $\pm$ 0.5\% &  3397.4ms $\pm$ 0.2\% & \\
+\hline
+\hline  crypto:       &   -                &  104.8ms $\pm$ 0.7\% &   104.7ms $\pm$ 0.4\% & \\
+\hline  deltablue:    &   -                &  206.7ms $\pm$ 0.4\% &   205.8ms $\pm$ 0.4\% & \\
+\hline  earley-boyer: &   1.009x as fast   & 1817.8ms $\pm$ 0.8\% &  1801.5ms $\pm$ 0.3\% &     significant \\
+\hline  raytrace:     &   -                &  585.9ms $\pm$ 0.5\% &   584.8ms $\pm$ 0.3\% & \\
+\hline  regexp:       &   -                &  492.9ms $\pm$ 0.6\% &   491.7ms $\pm$ 0.1\% & \\
+\hline  richards:     &   1.018x as slow   &  149.0ms $\pm$ 0.7\% &   151.7ms $\pm$ 0.3\% &     significant \\
+\hline  splay:        &   -                &   57.5ms $\pm$ 0.8\% &    57.1ms $\pm$ 0.6\% & \\
+\hline
+\end{tabular}
+\caption{v8bench benchmark results}
+\label{fig:v8bench}
+\end{figure}
+
+\begin{figure}[H]
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+    \textbf{Test}                   & \textbf{Comparison}       &     \textbf{From}         &        \textbf{To}           &  \textbf{Details} \\
+\hline\hline
+Total:                 & 1.001x as slow & 279.2ms $\pm$ 0.1\% &  279.6ms $\pm$ 0.1\%   &  significant \\
+\hline\hline
+3d:                  & 1.004x as slow &  52.5ms $\pm$ 0.3\% &   52.7ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{1em} cube:              & ??               &  25.1ms $\pm$ 0.3\% &   25.2ms $\pm$ 0.4\%   &  \textit{might} be 1.001x as slow \\
+\hline\hspace{1em} morph:             & ??               &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.003x as slow \\
+\hline\hspace{1em} raytrace:          & 1.009x as slow &  20.4ms $\pm$ 0.5\% &   20.6ms $\pm$ 0.4\%   &  significant \\
+\hline\hline
+access:              & 1.011x as fast   &  18.5ms $\pm$ 0.3\% &   18.3ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{1em} binary-trees:      & 1.019x as slow &   3.0ms $\pm$ 0.9\% &    3.0ms $\pm$ 0.9\%   &  significant \\
+\hline\hspace{1em} fannkuch:          & 1.034x as fast   &   8.8ms $\pm$ 0.3\% &    8.5ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} nbody:             & 1.035x as fast   &   3.6ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{1em} nsieve:            & 1.052x as slow &   3.1ms $\pm$ 0.6\% &    3.3ms $\pm$ 0.5\%   &  significant \\
+\hline\hline
+bitops:              & 1.008x as fast   &  11.4ms $\pm$ 0.4\% &   11.3ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} 3bit-bits-in-byte: & 1.009x as fast   &   1.0ms $\pm$ 0.6\% &    1.0ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{1em} bits-in-byte:      & 1.023x as fast   &   4.1ms $\pm$ 0.5\% &    4.0ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{1em} bitwise-and:       & ??               &   2.8ms $\pm$ 0.5\% &    2.8ms $\pm$ 0.6\%   &  \textit{might} be 1.005x as slow \\
+\hline\hspace{1em} nsieve-bits:       & -                &   3.6ms $\pm$ 0.5\% &    3.6ms $\pm$ 0.5\% & \\
+\hline\hline
+controlflow:         & 1.020x as slow &   2.3ms $\pm$ 0.7\% &    2.3ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} recursive:         & 1.020x as slow &   2.3ms $\pm$ 0.7\% &    2.3ms $\pm$ 0.4\%   &  significant \\
+\hline\hline
+crypto:              & 1.008x as slow &  27.3ms $\pm$ 0.3\% &   27.5ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{1em} aes:               & 1.009x as slow &  15.5ms $\pm$ 0.4\% &   15.6ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} md5:               & ??               &   6.9ms $\pm$ 0.7\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
+\hline\hspace{1em} sha1:              & 1.008x as slow &   4.9ms $\pm$ 0.5\% &    4.9ms $\pm$ 0.5\%   &  significant \\
+\hline\hline
+date:                & 1.008x as fast   &  62.6ms $\pm$ 0.2\% &   62.1ms $\pm$ 0.2\%   &  significant \\
+\hline\hspace{1em} format-tofte:      & -                &  42.6ms $\pm$ 0.2\% &   42.6ms $\pm$ 0.2\%   & \\
+\hline\hspace{1em} format-xparb:      & 1.023x as fast   &  20.0ms $\pm$ 0.3\% &   19.5ms $\pm$ 0.3\%   &  significant \\
+\hline\hline
+math:                & ??               &  13.2ms $\pm$ 0.3\% &   13.2ms $\pm$ 0.4\%   &  \textit{might} be 1.003x as slow \\
+\hline\hspace{1em} cordic:            & 1.021x as fast   &   2.7ms $\pm$ 0.4\% &    2.6ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{1em} partial-sums:      & -                &   7.1ms $\pm$ 0.5\% &    7.1ms $\pm$ 0.6\% & \\
+\hline\hspace{1em} spectral-norm:     & 1.024x as slow &   3.4ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.4\%   &  significant \\
+\hline\hline
+regexp:              & 1.012x as fast   &  10.2ms $\pm$ 0.5\% &   10.0ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} dna:               & 1.012x as fast   &  10.2ms $\pm$ 0.5\% &   10.0ms $\pm$ 0.4\%   &  significant \\
+\hline\hline
+string:              & 1.009x as slow &  81.3ms $\pm$ 0.2\% &   82.1ms $\pm$ 0.2\%   &  significant \\
+\hline\hspace{1em} base64:            & ??               &   4.9ms $\pm$ 0.6\% &    4.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
+\hline\hspace{1em} fasta:             & 1.012x as slow &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{1em} tagcloud:          & -                &  23.2ms $\pm$ 0.4\% &   23.3ms $\pm$ 0.3\% & \\
+\hline\hspace{1em} unpack-code:       & 1.014x as slow &  39.5ms $\pm$ 0.2\% &   40.1ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{1em} validate-input:    & 1.007x as slow &   6.8ms $\pm$ 0.4\% &    6.9ms $\pm$ 0.5\%   &  significant  \\
+\hline
+\end{tabular}
+\caption{SunSpider benchmark results}
+\label{fig:sunspider}
+\end{figure}
+
+\begin{figure}[H]
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+    \textbf{Test}     &\textbf{Comparison} & \textbf{From}        & \textbf{To}           &  \textbf{Details} \\
+\hline\hline
+    Total:                      & 1.015x as fast   & 2319.6ms $\pm$ 0.3\% &  2285.4ms $\pm$ 0.4\%  &   significant\\
+\hline\hline
+      ai:                       & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
+\hline\hspace{1em} astar:                  & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
+\hline\hline
+      audio:                    & -                &  842.8ms $\pm$ 0.4\% &   838.0ms $\pm$ 0.7\%  &\\
+\hline\hspace{1em} beat-detection:         & ??               &  393.0ms $\pm$ 1.0\% &   397.6ms $\pm$ 1.5\%  &   \textit{might} be 1.012x as slow\\
+\hline\hspace{1em} dft:                    & 1.030x as fast   &  162.6ms $\pm$ 0.6\% &   157.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{1em} fft:                    & 1.016x as fast   &  123.5ms $\pm$ 0.2\% &   121.6ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{1em} oscillator:             & 1.017x as fast   &  163.7ms $\pm$ 0.3\% &   160.9ms $\pm$ 0.3\%  &   significant\\
+\hline\hline
+      imaging:                  & 1.020x as fast   &  569.1ms $\pm$ 0.9\% &   557.9ms $\pm$ 1.3\%  &   significant\\
+\hline\hspace{1em} gaussian-blur:          & -                &  197.2ms $\pm$ 1.7\% &   193.7ms $\pm$ 2.1\%  &\\
+\hline\hspace{1em} darkroom:               & -                &  248.2ms $\pm$ 0.3\% &   247.6ms $\pm$ 0.4\%  &\\
+\hline\hspace{1em} desaturate:             & 1.062x as fast   &  123.8ms $\pm$ 3.2\% &   116.7ms $\pm$ 5.1\%  &   significant\\
+\hline\hline
+      json:                     & 1.019x as fast   &  121.3ms $\pm$ 0.2\% &   119.0ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{1em} parse-financial:        & 1.005x as fast   &   75.8ms $\pm$ 0.2\% &    75.5ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{1em} stringify-tinderbox:    & 1.044x as fast   &   45.5ms $\pm$ 0.5\% &    43.5ms $\pm$ 0.8\%  &   significant\\
+\hline\hline
+      stanford:                 & -                &  665.0ms $\pm$ 0.5\% &   662.2ms $\pm$ 0.4\%  &\\
+\hline\hspace{1em} crypto-aes:             & -                &  160.7ms $\pm$ 1.5\% &   160.6ms $\pm$ 1.1\%  &\\
+\hline\hspace{1em} crypto-ccm:             & 1.008x as slow &  158.6ms $\pm$ 0.4\% &   159.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{1em} crypto-pbkdf2:          & 1.006x as fast   &  254.9ms $\pm$ 0.5\% &   253.3ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{1em} crypto-sha256-iterative:& 1.025x as fast   &   90.8ms $\pm$ 0.4\% &    88.5ms $\pm$ 0.5\%  &   significant \\
+\hline
+\end{tabular}
+\caption{Kraken benchmark results}
+\label{fig:kraken}
+\end{figure}
 %% XXX: TODO

### 2012-05-02 project-writings/paper.tex
summary and more words

@@
 are consulted to determine whether to perform overflow checks for
-arithmetic computations.
-
-OTHER OPTIMIZATIONS WE DO??
+arithmetic computations.  In addition to eliminating overflow checks, in
+limited cases we also eliminate some bounds checks from arrays.
 
@@
 \end{figure}
-%% XXX: TODO
+
+To summarize our evaluation, we discovered the checks we eliminate do result in
+some measurable performance increase. In extremly short-running tests such
+Sunspider (see Figure \ref{fig:sunspider}), the additional compile-time
+overhead that our pass requires yields slightly slower performance.
+
+%% XXX TODO MORE HERE
 
@@
 %% XXX: TODO
+%% Were any of your results surprising our counter-intuitive? If so, what did you learn from those surprises? What did you learn from this experience in general?
+Boy, JS is a pain in the ass.
+This is a lesson learned, not a surprise.
 
-\section{Conclusions}
-%% XXX: TODO
-Boy, JS is a pain in the ass.
+\section{Future Work and Conclusions}
+We implement an SSA-based range analysis pass which is capable of inferring
+range information on integer operations. We use the results of this analysis to
+perform some simple optimizations relevant to Javascript: we eliminate overflow
+checks which can be proven redundant, and also eliminate a limited number of
+array bounds checks.
+
+Further optimizations are possible using our range analysis framework. If the
+range associated with an instruction becomes empty, then the block has been
+proven dead and can be eliminated. Control flow operations can be proven
+redundant as well, further eliminating computation and dead basic blocks. These
+graph manipulations are currently not available in the IonMonkey compiler,
+which is in active development but not yet complete.
+
+Symbolic range propagation, where the bounds are expressed in terms of
+definitions rather than integer constants, is left as future work. Symbolic
+range propagation would allow more sophisticated removal of bounds checks,
+which are generally not based on constants.
+
+In summary, we augment the IonMonkey compiler with range information and begin
+to use this information for the purposes of optimization. We leave more
+sophisticated uses of the analysis as future work.
 

### 2012-05-02 project-writings/paper.tex
s/javascript/JavaScript/ig

@@
 With the explosion of browser based applications that has occurred in
-recent years, Javascript has become one of the most widely used
-programming languages, and Javascript performance has become
-critically important. Unfortunately, Javascript seems to be designed
+recent years, JavaScript has become one of the most widely used
+programming languages, and JavaScript performance has become
+critically important. Unfortunately, JavaScript seems to be designed
 to be un-performant.
 
-One source of these woes are Javascript's dynamic semantics for
+One source of these woes are JavaScript's dynamic semantics for
 numbers: all numeric values are defined by the specification to be
@@
 values are integers and because floating point operations are
-more expensive, all major Javascript implementations attempt to
+more expensive, all major JavaScript implementations attempt to
 represent numbers as integers when possible.
 
-In IonMonkey, Mozilla's next-generation optimizing Javascript JIT
+In IonMonkey, Mozilla's next-generation optimizing JavaScript JIT
 compiler, a hybrid type inference system (which uses both static analysis and
@@
 imperative SSA compiler. IonMonkey takes the bytecode used by the SpiderMonkey
-Javascript interpreter and compiles it to native code. Optimistic type
+JavaScript interpreter and compiles it to native code. Optimistic type
 specialization is used, and if type assumptions fail, then execution ``bails
@@
 
-While the ECMA specification for Javascript requires that all numbers behave as
+While the ECMA specification for JavaScript requires that all numbers behave as
 double-precision floating point values\cite{ECMA-262}, for performance reasons JIT compilers
@@
 does not use Cytron et al.'s algorithm for computing SSA form; it
-instead takes advantage of the structured nature of Javascript code to
+instead takes advantage of the structured nature of JavaScript code to
 go to SSA form without computing dominance frontiers \footnote{More
@@
 
-There are three benchmark suites which are generally used to measure Javascript
+There are three benchmark suites which are generally used to measure JavaScript
 JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
@@
 range information on integer operations. We use the results of this analysis to
-perform some simple optimizations relevant to Javascript: we eliminate overflow
+perform some simple optimizations relevant to JavaScript: we eliminate overflow
 checks which can be proven redundant, and also eliminate a limited number of

### 2012-05-02 project-writings/paper.tex
Equal contribution to this work
cite more things

@@
 \author{
-Ryan Pearl\\
+Ryan Pearl\footnote{Equal contribution to this work}\\
 \texttt{$<$rpearl@andrew.cmu.edu$>$}\\
 \and
-Michael Sullivan\\
+Michael Sullivan\footnotemark[\value{footnote}]\\
 \texttt{$<$mjsulliv@cs.cmu.edu$>$}\\
@@
 workflow algorithm due to Patterson in ``Accurate Static Branch Prediction by
-Value Range Propagation.'' The orignal non-SSA based dataflow algorithm is by
-Harrison in ``Compiler Analysis of the Value Ranges of Variables.''
+Value Range Propagation.''\cite{Patterson95accuratestatic}
+The orignal non-SSA based dataflow algorithm is by
+Harrison in ``Compiler Analysis of the Value Ranges of Variables.''\cite{Harrison}
 

### 2012-05-02 project-writings/paper.tex
grammar nits

@@
   structures of a modern SSA compiler.
-\item To facilitate the above, extended IonMonkey with a fast
+\item To facilitate the above, extending IonMonkey with a fast
   (constant time, one integer per basic block memory overhead) 
@@
 The IonMonkey compiler is somewhat loosely based on LLVM, in that it is an
-imperative SSA compiler. IonMonkey takes the bytecode used by the SpiderMonkey
-JavaScript interpreter and compiles it to native code. Optimistic type
-specialization is used, and if type assumptions fail, then execution ``bails
-out'' to the interpreter.
+imperative SSA-based compiler. IonMonkey takes the bytecode used by the SpiderMonkey
+JavaScript interpreter and compiles it to native code. IonMonkey uses optimistic type
+specialization, and if type assumptions fail, then execution ``bails
+out'' to the interpreter. Bailout points are inserted when global effects take place.
 

### 2012-05-02 project-writings/paper.tex
'selected' range analysis equations since we also do things like charCodeAt and >>, << and shit

@@
 
-\caption{Range arithmetic definitions}
+\caption{Selected range arithmetic definitions}
 \label{fig:range_arith}
@@
 \end{eqnarray*}
-\caption{Range analysis equations}
+\caption{Selected range analysis equations}
 \label{fig:range_anal}
@@
 analysis pass, and measured the performance differences. The results are
-summarized below:
+summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and \ref{fig:kraken}.
 
@@
 
+In order to get a different view of how well our optimization was
+performing, we annotated IonMonkey to emit for each addition
+instruction compiled whether or not range analysis proved that it
+could not overflow, and ran this against all of the tests in Kraken,
+Sunspider, and v8bench. The data from these tests are presented in Figure
+\ref{fig:count} (tests that were found to have very few total add
+instructions were not included).
+
+We ran this experiment in two configurations: with and without
+on-stack replacement (OSR) enabled. On-stack replacement is a
+mechanism by which IonMonkey can generate code that allows entering a
+function at a point other than the start; specifically, it allows
+entering in the middle of a loop. OSR allows IonMonkey to switch to
+JITted code in the middle of a loop that is discovered to be hot,
+instead of needing to wait for the function it is in to finish
+first. OSR's impact on how many overlow checks we can eliminate is
+somewhat unpredictable. OSR being on introduces new edges to the
+control flow graph that can result in less precise range
+information. However, OSR also causes more code to be JITted,
+increasing potential opportunities to find ranges. Both of these
+effects come into play in different tests.
+
+These results show that our range analysis is generally quite
+successful in eliminating overflow checks from programs, but our
+timing results indicate that this does yield that substantial of a
+performance improvement. We believe that this is because the overhead
+of an essentially never taken branch on modern hardware are quite
+low. While on a microbenchmark we constructed overflow check
+elimination resulted in a substantial (around 30\%) speedup, it appears that
+this does not translate well to more realistic code.
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=0.98\textwidth]{counts_graph.png}
+\caption{Percentage of add overflow checks eliminated}
+\label{fig:count}
+\end{figure*}
+
 \section{Surprises and Lessons Learned}

### 2012-05-02 project-writings/paper.tex
charts

--- 2012-05-02 project-writings/paper.tex
@@
 \author{
-Ryan Pearl\footnote{Equal contribution to this work}\\
+Ryan Pearl\\
 \texttt{$<$rpearl@andrew.cmu.edu$>$}\\
 \and
-Michael Sullivan\footnotemark[\value{footnote}]\\
+Michael Sullivan\\
 \texttt{$<$mjsulliv@cs.cmu.edu$>$}\\
@@
 workflow algorithm due to Patterson in ``Accurate Static Branch Prediction by
-Value Range Propagation.''\cite{Patterson95accuratestatic}
-The orignal non-SSA based dataflow algorithm is by
-Harrison in ``Compiler Analysis of the Value Ranges of Variables.''\cite{Harrison}
+Value Range Propagation.'' The orignal non-SSA based dataflow algorithm is by
+Harrison in ``Compiler Analysis of the Value Ranges of Variables.''
 
@@
 
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=0.98\textwidth]{counts_graph.png}
+\caption{Percentage of add overflow checks eliminated}
+\label{fig:count}
+\end{figure*}
+
 \section{Surprises and Lessons Learned}

### 2012-05-02 project-writings/count-graph.py

@@ -1,0 +1,119 @@
+#!/usr/bin/env python
+
+import sys
+from scipy.stats import stats
+import Gnuplot, Gnuplot.funcutils
+
+measurements = ["Add"]
+
+def process_file(file_name):
+    bs, name, mode, suite = file_name.split("_")
+    name = name.replace("check-", "")
+    name = name.replace("run-", "")
+    suite = "test" # HACK
+    lines = [line.strip() for line in open(file_name)]
+    counts = {}
+    for s in measurements:
+        total, finite, truncated, both = 0, 0, 0, 0
+        for line in lines:
+            if s in line:
+                _, _, is_finite, is_truncated = line.split(" ")
+                is_finite, is_truncated = int(is_finite), int(is_truncated)
+                if is_finite: finite += 1
+                if is_truncated: truncated += 1
+                if is_finite and is_truncated: both += 1
+                total += 1
+        counts[s] = (total, finite, truncated, both)
+                
+    return (name, suite, mode, counts)
+
+def map_add(m, key, x):
+    if key not in m:
+        m[key] = []
+    m[key].append(x)
+
+def organize_data(data, suite):
+    data = [x[:1]+x[2:] for x in data if x[1] == suite]
+    modes = {(x[1], y): [] for x in data for y in measurements}
+    for x in data:
+        for k, v in x[2].iteritems():
+            #if v[0] == 0: v = 1
+            #else: v = float(v[1])/float(v[0])
+            modes[(x[1], k)].append((x[0], (v[0], v[1])))
+    for v in modes.values():
+        v.sort()
+    return modes
+
+def reorganize_data(data, suite):
+    modes = list(data.keys())
+    data2 = {}
+    for k, v in data.iteritems():
+        for (test, entry) in v:
+            map_add(data2, test, entry)
+    return (modes, data2)
+
+def produce_stats(data):
+    m = {}
+    for (server, size, conns, it, time, rate) in data:
+        map_add(m, (server, size, conns), rate)
+    data = []
+    for k, v in m.items():
+        mean = stats.nanmean(v)
+        stddev = stats.nanstd(v)
+        data += [k + (mean, stddev)]
+    return data
+
+def spew_data(suite, table):
+    modes, table = table
+    f = open(suite + "_counts.csv", "w")
+    i = 0
+    for test, points in table.iteritems():
+        if all(x[0] < 5 for x in points): continue
+        f.write(test + ',')
[…]

### 2012-05-02 project-writings/paper.tex
typo

--- 2012-05-02 project-writings/paper.tex
@@
 Production Compiler'', by Teixeira and Pereira, outlines a practical implementation of
-implementation of an algorithm designed by by Su and Wagner
+implementation of an algorithm designed by Su and Wagner
 \cite{Su04aclass}. The algorithm due to Su and Wagner is

### 2012-05-02 project-writings/paper.tex
A bunch more evaluation.

--- 2012-05-02 project-writings/paper.tex
@@
 
-\caption{Selected range arithmetic definitions}
+\caption{Range arithmetic definitions}
 \label{fig:range_arith}
@@
 \end{eqnarray*}
-\caption{Selected range analysis equations}
+\caption{Range analysis equations}
 \label{fig:range_anal}
@@
 low. While on a microbenchmark we constructed overflow check
-elimination resulted in a substantial (around 30\%) speedup, it appears that
+elimination resulted in a substantial (!30\%) speedup, it appears that
 this does not translate well to more realistic code.

### 2012-05-02 project-writings/paper.tex
oops

@@
 low. While on a microbenchmark we constructed overflow check
-elimination resulted in a substantial (!30\%) speedup, it appears that
+elimination resulted in a substantial (around 30\%) speedup, it appears that
 this does not translate well to more realistic code.

### 2012-05-02 project-writings/paper.tex
bells and whistles

--- 2012-05-02 project-writings/paper.tex
@@
 \usepackage{listings}
+\usepackage{color}
+\definecolor{lightgray}{rgb}{.9,.9,.9}
 %\usepackage[all]{hypcap}
@@
   language=Python,
+  backgroundcolor=\color{lightgray},
   showstringspaces=false,
@@
   captionpos=b
+}
+\lstdefinelanguage{JavaScript}{
+  keywords={if,else,var},
+  keywordstyle=\bfseries\ttfamily,
+  backgroundcolor=\color{lightgray},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/}
 }
@@
 Intermediate Representation'' (MIR) which encompasses higher-level
-constructs such as lambdas, objects, and strings. A variety of
-optimizations are performed, and then the code is lowered to a ``Lower
-Intermediate Representation'' (LIR) used for register allocation and
-code generation.
+constructs such as lambdas, objects, and strings. IonMonkey performs a variety of
+optimizations including global value numbering and loop invariant code motion,
+and then the code is lowered to a ``Lower Intermediate Representation'' (LIR)
+used for register allocation and code generation. IonMonkey currently supports
+code generation for the x86, x86\_64, and ARM architectures.
 
@@
 
-\begin{figure}[ht]
-\begin{center}
-\begin{verbatim}
+\begin{lstlisting}[language=Javascript,
+                   caption={Code that requires information about branches to properly analyze},
+                   label={fig:branch}, float=ht]
 var y;
@@
 }
-\end{verbatim}
-\end{center}
-\caption{Code that requires information about branches to properly analyze}
-\label{fig:branch}
-\end{figure}
+\end{lstlisting}
 
-\begin{figure}[ht]
-\begin{center}
-\begin{verbatim}
+\begin{lstlisting}[language=Javascript,
+                   caption={Code from Figure \ref{fig:branch}, in XSA form},
+                   label={fig:xsa}, float=ht]
 if (x < 0) {
@@
 y = Phi(y1, y4);
-\end{verbatim}
-\end{center}
-\caption{Code from Figure \ref{fig:branch}, in XSA form}
-\label{fig:xsa}
-\end{figure}
+\end{lstlisting}
 
@@
                    caption={Pseudocode algorithm for inserting beta nodes},
-                   label={lst:add_betas}]
+                   label={lst:add_betas}, float=ht]
 def addBetaNodesToFunc(cfg):
@@
                    caption={Pseudocode algorithm for propagating range information},
-                   label={lst:propagate_ranges}]
+                   label={lst:propagate_ranges}, float=ht]
 def propagateRanges(graph):

### 2012-05-02 project-writings/paper.tex
cite the benchmarks and stats

@@
 There are three benchmark suites which are generally used to measure JavaScript
-JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
-suite developed by Apple, and the Kraken benchmark suite developed by Mozilla.
+JIT compilers: the v8bench suite developed by Google\cite{v8bench}, the SunSpider benchmark
+suite developed by Apple\cite{sunspider}, and the Kraken benchmark suite developed by Mozilla\cite{kraken}
 
@@
 analysis pass, and measured the performance differences. The results are
-summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and \ref{fig:kraken}.
+summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and
+\ref{fig:kraken}. These figures were gathered using the SunSpider test harness,
+which gives both a mean and a 95\% confidence interval\cite{sunspider}. Each
+test was run 30 times.
 
@@
 \section{Future Work and Conclusions}
-We implement an SSA-based range analysis pass which is capable of inferring
-range information on integer operations. We use the results of this analysis to
-perform some simple optimizations relevant to JavaScript: we eliminate overflow
-checks which can be proven redundant, and also eliminate a limited number of
-array bounds checks.
+We implement an SSA-based range analysis pass which is capable of
+inferring range information on integer operations. We use the results
+of this analysis to perform some simple optimizations relevant to
+JavaScript: we eliminate overflow checks which can be proven
+redundant, and also eliminate a limited number of array bounds checks.
 
-Further optimizations are possible using our range analysis framework. If the
-range associated with an instruction becomes empty, then the block has been
-proven dead and can be eliminated. Control flow operations can be proven
-redundant as well, further eliminating computation and dead basic blocks. These
-graph manipulations are currently not available in the IonMonkey compiler,
+On a suite of standard Javascript benchmarks, our range analysis
+succeeds in eliminating a large fraction of overflow checks for
+integer additions. This translates into a minor overall performance
+win on the Kraken benchmark suite and a very minor overall performance
+loss on the Sunspider suite. Since we should only increase the
+quality of the generated code, any performance losses should be from
+increased compilation overhead (Sunspider tests in particular are
+quite short). The performance improvements are decent, but not
+stellar; we believe this is because Javascript code is generally not
+integer-arithmetic-heavy tight loops and because the overflow test
+branches are generally predicted correctly.
+
+A number of further optimizations are possible using our range
+analysis framework. If the range associated with an instruction
+becomes empty, then the block has been proven dead and can be
+eliminated. Control flow operations can be proven redundant as well,
+further eliminating computation and dead basic blocks. These graph
+manipulations are currently not available in the IonMonkey compiler,
 which is in active development but not yet complete.
@@
 Symbolic range propagation, where the bounds are expressed in terms of
-definitions rather than integer constants, is left as future work. Symbolic
-range propagation would allow more sophisticated removal of bounds checks,
-which are generally not based on constants.
+definitions rather than integer constants, is left as future
+work. Symbolic range propagation would allow more sophisticated
+removal of bounds checks, which are generally not based on
+constants. It would also allow for much more potential for eliminating
+redundant branches than just constant ranges would.
 
-In summary, we augment the IonMonkey compiler with range information and begin
-to use this information for the purposes of optimization. We leave more
-sophisticated uses of the analysis as future work.
+We believe that we have implemented a worthwhile optimization and a
+useful analysis pass that will serve as a good building block for
+future IonMonkey optimizations. Our range analysis patches are likely
+to land in IonMonkey in the next month, and it is likely that some of
+the future work we discuss will be taken up by the IonMonkey team.
 

### 2012-05-02 project-writings/paper.tex
aesthetics of tables

@@
 3d:                  & 1.004x as slow &  52.5ms $\pm$ 0.3\% &   52.7ms $\pm$ 0.3\%   &  significant \\
-\hline\hspace{1em} cube:              & ??               &  25.1ms $\pm$ 0.3\% &   25.2ms $\pm$ 0.4\%   &  \textit{might} be 1.001x as slow \\
-\hline\hspace{1em} morph:             & ??               &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.003x as slow \\
-\hline\hspace{1em} raytrace:          & 1.009x as slow &  20.4ms $\pm$ 0.5\% &   20.6ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} cube:              & ??               &  25.1ms $\pm$ 0.3\% &   25.2ms $\pm$ 0.4\%   &  \textit{might} be 1.001x as slow \\
+\hline\hspace{0.5em} morph:             & ??               &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.003x as slow \\
+\hline\hspace{0.5em} raytrace:          & 1.009x as slow &  20.4ms $\pm$ 0.5\% &   20.6ms $\pm$ 0.4\%   &  significant \\
 \hline\hline
 access:              & 1.011x as fast   &  18.5ms $\pm$ 0.3\% &   18.3ms $\pm$ 0.3\%   &  significant \\
-\hline\hspace{1em} binary-trees:      & 1.019x as slow &   3.0ms $\pm$ 0.9\% &    3.0ms $\pm$ 0.9\%   &  significant \\
-\hline\hspace{1em} fannkuch:          & 1.034x as fast   &   8.8ms $\pm$ 0.3\% &    8.5ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} nbody:             & 1.035x as fast   &   3.6ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.3\%   &  significant \\
-\hline\hspace{1em} nsieve:            & 1.052x as slow &   3.1ms $\pm$ 0.6\% &    3.3ms $\pm$ 0.5\%   &  significant \\
+\hline\hspace{0.5em} binary-trees:      & 1.019x as slow &   3.0ms $\pm$ 0.9\% &    3.0ms $\pm$ 0.9\%   &  significant \\
+\hline\hspace{0.5em} fannkuch:          & 1.034x as fast   &   8.8ms $\pm$ 0.3\% &    8.5ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} nbody:             & 1.035x as fast   &   3.6ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{0.5em} nsieve:            & 1.052x as slow &   3.1ms $\pm$ 0.6\% &    3.3ms $\pm$ 0.5\%   &  significant \\
 \hline\hline
 bitops:              & 1.008x as fast   &  11.4ms $\pm$ 0.4\% &   11.3ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} 3bit-bits-in-byte: & 1.009x as fast   &   1.0ms $\pm$ 0.6\% &    1.0ms $\pm$ 0.6\%   &  significant \\
-\hline\hspace{1em} bits-in-byte:      & 1.023x as fast   &   4.1ms $\pm$ 0.5\% &    4.0ms $\pm$ 0.6\%   &  significant \\
-\hline\hspace{1em} bitwise-and:       & ??               &   2.8ms $\pm$ 0.5\% &    2.8ms $\pm$ 0.6\%   &  \textit{might} be 1.005x as slow \\
-\hline\hspace{1em} nsieve-bits:       & -                &   3.6ms $\pm$ 0.5\% &    3.6ms $\pm$ 0.5\% & \\
+\hline\hspace{0.5em} 3bit-bits-in-byte: & 1.009x as fast   &   1.0ms $\pm$ 0.6\% &    1.0ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{0.5em} bits-in-byte:      & 1.023x as fast   &   4.1ms $\pm$ 0.5\% &    4.0ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{0.5em} bitwise-and:       & ??               &   2.8ms $\pm$ 0.5\% &    2.8ms $\pm$ 0.6\%   &  \textit{might} be 1.005x as slow \\
+\hline\hspace{0.5em} nsieve-bits:       & -                &   3.6ms $\pm$ 0.5\% &    3.6ms $\pm$ 0.5\% & \\
 \hline\hline
 controlflow:         & 1.020x as slow &   2.3ms $\pm$ 0.7\% &    2.3ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} recursive:         & 1.020x as slow &   2.3ms $\pm$ 0.7\% &    2.3ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} recursive:         & 1.020x as slow &   2.3ms $\pm$ 0.7\% &    2.3ms $\pm$ 0.4\%   &  significant \\
 \hline\hline
 crypto:              & 1.008x as slow &  27.3ms $\pm$ 0.3\% &   27.5ms $\pm$ 0.3\%   &  significant \\
-\hline\hspace{1em} aes:               & 1.009x as slow &  15.5ms $\pm$ 0.4\% &   15.6ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} md5:               & ??               &   6.9ms $\pm$ 0.7\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
-\hline\hspace{1em} sha1:              & 1.008x as slow &   4.9ms $\pm$ 0.5\% &    4.9ms $\pm$ 0.5\%   &  significant \\
+\hline\hspace{0.5em} aes:               & 1.009x as slow &  15.5ms $\pm$ 0.4\% &   15.6ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} md5:               & ??               &   6.9ms $\pm$ 0.7\% &    6.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
+\hline\hspace{0.5em} sha1:              & 1.008x as slow &   4.9ms $\pm$ 0.5\% &    4.9ms $\pm$ 0.5\%   &  significant \\
 \hline\hline
 date:                & 1.008x as fast   &  62.6ms $\pm$ 0.2\% &   62.1ms $\pm$ 0.2\%   &  significant \\
-\hline\hspace{1em} format-tofte:      & -                &  42.6ms $\pm$ 0.2\% &   42.6ms $\pm$ 0.2\%   & \\
-\hline\hspace{1em} format-xparb:      & 1.023x as fast   &  20.0ms $\pm$ 0.3\% &   19.5ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{0.5em} format-tofte:      & -                &  42.6ms $\pm$ 0.2\% &   42.6ms $\pm$ 0.2\%   & \\
+\hline\hspace{0.5em} format-xparb:      & 1.023x as fast   &  20.0ms $\pm$ 0.3\% &   19.5ms $\pm$ 0.3\%   &  significant \\
 \hline\hline
 math:                & ??               &  13.2ms $\pm$ 0.3\% &   13.2ms $\pm$ 0.4\%   &  \textit{might} be 1.003x as slow \\
-\hline\hspace{1em} cordic:            & 1.021x as fast   &   2.7ms $\pm$ 0.4\% &    2.6ms $\pm$ 0.6\%   &  significant \\
-\hline\hspace{1em} partial-sums:      & -                &   7.1ms $\pm$ 0.5\% &    7.1ms $\pm$ 0.6\% & \\
-\hline\hspace{1em} spectral-norm:     & 1.024x as slow &   3.4ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} cordic:            & 1.021x as fast   &   2.7ms $\pm$ 0.4\% &    2.6ms $\pm$ 0.6\%   &  significant \\
+\hline\hspace{0.5em} partial-sums:      & -                &   7.1ms $\pm$ 0.5\% &    7.1ms $\pm$ 0.6\% & \\
+\hline\hspace{0.5em} spectral-norm:     & 1.024x as slow &   3.4ms $\pm$ 0.6\% &    3.5ms $\pm$ 0.4\%   &  significant \\
 \hline\hline
 regexp:              & 1.012x as fast   &  10.2ms $\pm$ 0.5\% &   10.0ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} dna:               & 1.012x as fast   &  10.2ms $\pm$ 0.5\% &   10.0ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} dna:               & 1.012x as fast   &  10.2ms $\pm$ 0.5\% &   10.0ms $\pm$ 0.4\%   &  significant \\
 \hline\hline
 string:              & 1.009x as slow &  81.3ms $\pm$ 0.2\% &   82.1ms $\pm$ 0.2\%   &  significant \\
-\hline\hspace{1em} base64:            & ??               &   4.9ms $\pm$ 0.6\% &    4.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
-\hline\hspace{1em} fasta:             & 1.012x as slow &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  significant \\
-\hline\hspace{1em} tagcloud:          & -                &  23.2ms $\pm$ 0.4\% &   23.3ms $\pm$ 0.3\% & \\
-\hline\hspace{1em} unpack-code:       & 1.014x as slow &  39.5ms $\pm$ 0.2\% &   40.1ms $\pm$ 0.3\%   &  significant \\
-\hline\hspace{1em} validate-input:    & 1.007x as slow &   6.8ms $\pm$ 0.4\% &    6.9ms $\pm$ 0.5\%   &  significant  \\
+\hline\hspace{0.5em} base64:            & ??               &   4.9ms $\pm$ 0.6\% &    4.9ms $\pm$ 0.4\%   &  \textit{might} be 1.006x as slow \\
+\hline\hspace{0.5em} fasta:             & 1.012x as slow &   6.9ms $\pm$ 0.5\% &    6.9ms $\pm$ 0.4\%   &  significant \\
+\hline\hspace{0.5em} tagcloud:          & -                &  23.2ms $\pm$ 0.4\% &   23.3ms $\pm$ 0.3\% & \\
+\hline\hspace{0.5em} unpack-code:       & 1.014x as slow &  39.5ms $\pm$ 0.2\% &   40.1ms $\pm$ 0.3\%   &  significant \\
+\hline\hspace{0.5em} validate-input:    & 1.007x as slow &   6.8ms $\pm$ 0.4\% &    6.9ms $\pm$ 0.5\%   &  significant  \\
 \hline
@@
       ai:                       & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
-\hline\hspace{1em} astar:                  & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
+\hline\hspace{0.5em} astar:                  & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
 \hline\hline
       audio:                    & -                &  842.8ms $\pm$ 0.4\% &   838.0ms $\pm$ 0.7\%  &\\
-\hline\hspace{1em} beat-detection:         & ??               &  393.0ms $\pm$ 1.0\% &   397.6ms $\pm$ 1.5\%  &   \textit{might} be 1.012x as slow\\
-\hline\hspace{1em} dft:                    & 1.030x as fast   &  162.6ms $\pm$ 0.6\% &   157.8ms $\pm$ 0.5\%  &   significant\\
-\hline\hspace{1em} fft:                    & 1.016x as fast   &  123.5ms $\pm$ 0.2\% &   121.6ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{1em} oscillator:             & 1.017x as fast   &  163.7ms $\pm$ 0.3\% &   160.9ms $\pm$ 0.3\%  &   significant\\
+\hline\hspace{0.5em} beat-detection:         & ??               &  393.0ms $\pm$ 1.0\% &   397.6ms $\pm$ 1.5\%  &   \textit{might} be 1.012x as slow\\
+\hline\hspace{0.5em} dft:                    & 1.030x as fast   &  162.6ms $\pm$ 0.6\% &   157.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{0.5em} fft:                    & 1.016x as fast   &  123.5ms $\pm$ 0.2\% &   121.6ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} oscillator:             & 1.017x as fast   &  163.7ms $\pm$ 0.3\% &   160.9ms $\pm$ 0.3\%  &   significant\\
 \hline\hline
       imaging:                  & 1.020x as fast   &  569.1ms $\pm$ 0.9\% &   557.9ms $\pm$ 1.3\%  &   significant\\
-\hline\hspace{1em} gaussian-blur:          & -                &  197.2ms $\pm$ 1.7\% &   193.7ms $\pm$ 2.1\%  &\\
-\hline\hspace{1em} darkroom:               & -                &  248.2ms $\pm$ 0.3\% &   247.6ms $\pm$ 0.4\%  &\\
-\hline\hspace{1em} desaturate:             & 1.062x as fast   &  123.8ms $\pm$ 3.2\% &   116.7ms $\pm$ 5.1\%  &   significant\\
+\hline\hspace{0.5em} gaussian-blur:          & -                &  197.2ms $\pm$ 1.7\% &   193.7ms $\pm$ 2.1\%  &\\
+\hline\hspace{0.5em} darkroom:               & -                &  248.2ms $\pm$ 0.3\% &   247.6ms $\pm$ 0.4\%  &\\
+\hline\hspace{0.5em} desaturate:             & 1.062x as fast   &  123.8ms $\pm$ 3.2\% &   116.7ms $\pm$ 5.1\%  &   significant\\
 \hline\hline
       json:                     & 1.019x as fast   &  121.3ms $\pm$ 0.2\% &   119.0ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{1em} parse-financial:        & 1.005x as fast   &   75.8ms $\pm$ 0.2\% &    75.5ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{1em} stringify-tinderbox:    & 1.044x as fast   &   45.5ms $\pm$ 0.5\% &    43.5ms $\pm$ 0.8\%  &   significant\\
+\hline\hspace{0.5em} parse-financial:        & 1.005x as fast   &   75.8ms $\pm$ 0.2\% &    75.5ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} stringify-tinderbox:    & 1.044x as fast   &   45.5ms $\pm$ 0.5\% &    43.5ms $\pm$ 0.8\%  &   significant\\
 \hline\hline
       stanford:                 & -                &  665.0ms $\pm$ 0.5\% &   662.2ms $\pm$ 0.4\%  &\\
-\hline\hspace{1em} crypto-aes:             & -                &  160.7ms $\pm$ 1.5\% &   160.6ms $\pm$ 1.1\%  &\\
-\hline\hspace{1em} crypto-ccm:             & 1.008x as slow &  158.6ms $\pm$ 0.4\% &   159.8ms $\pm$ 0.5\%  &   significant\\
-\hline\hspace{1em} crypto-pbkdf2:          & 1.006x as fast   &  254.9ms $\pm$ 0.5\% &   253.3ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{1em} crypto-sha256-iterative:& 1.025x as fast   &   90.8ms $\pm$ 0.4\% &    88.5ms $\pm$ 0.5\%  &   significant \\
+\hline\hspace{0.5em} crypto-aes:             & -                &  160.7ms $\pm$ 1.5\% &   160.6ms $\pm$ 1.1\%  &\\
+\hline\hspace{0.5em} crypto-ccm:             & 1.008x as slow &  158.6ms $\pm$ 0.4\% &   159.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{0.5em} crypto-pbkdf2:          & 1.006x as fast   &  254.9ms $\pm$ 0.5\% &   253.3ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} crypto-sha256-iterative:& 1.025x as fast   &   90.8ms $\pm$ 0.4\% &    88.5ms $\pm$ 0.5\%  &   significant \\
 \hline

### 2012-05-02 project-writings/paper.tex
remove the Surprises and Lessons Learned section
Do a bunch of work on the conclusion.

--- 2012-05-02 project-writings/paper.tex
@@
 \usepackage{listings}
-\usepackage{color}
-\definecolor{lightgray}{rgb}{.9,.9,.9}
 %\usepackage[all]{hypcap}
@@
   language=Python,
-  backgroundcolor=\color{lightgray},
   showstringspaces=false,
@@
   captionpos=b
-}
-\lstdefinelanguage{JavaScript}{
-  keywords={if,else,var},
-  keywordstyle=\bfseries\ttfamily,
-  backgroundcolor=\color{lightgray},
-  sensitive=false,
-  comment=[l]{//},
-  morecomment=[s]{/*}{*/}
 }
@@
 Production Compiler'', by Teixeira and Pereira, outlines a practical implementation of
-implementation of an algorithm designed by Su and Wagner
+implementation of an algorithm designed by by Su and Wagner
 \cite{Su04aclass}. The algorithm due to Su and Wagner is
@@
 Intermediate Representation'' (MIR) which encompasses higher-level
-constructs such as lambdas, objects, and strings. IonMonkey performs a variety of
-optimizations including global value numbering and loop invariant code motion,
-and then the code is lowered to a ``Lower Intermediate Representation'' (LIR)
-used for register allocation and code generation. IonMonkey currently supports
-code generation for the x86, x86\_64, and ARM architectures.
+constructs such as lambdas, objects, and strings. A variety of
+optimizations are performed, and then the code is lowered to a ``Lower
+Intermediate Representation'' (LIR) used for register allocation and
+code generation.
 
@@
 
-\begin{lstlisting}[language=Javascript,
-                   caption={Code that requires information about branches to properly analyze},
-                   label={fig:branch}, float=ht]
+\begin{figure}[ht]
+\begin{center}
+\begin{verbatim}
 var y;
@@
 }
-\end{lstlisting}
+\end{verbatim}
+\end{center}
+\caption{Code that requires information about branches to properly analyze}
+\label{fig:branch}
+\end{figure}
 
-\begin{lstlisting}[language=Javascript,
-                   caption={Code from Figure \ref{fig:branch}, in XSA form},
-                   label={fig:xsa}, float=ht]
+\begin{figure}[ht]
+\begin{center}
+\begin{verbatim}
 if (x < 0) {
@@
 y = Phi(y1, y4);
-\end{lstlisting}
+\end{verbatim}
+\end{center}
+\caption{Code from Figure \ref{fig:branch}, in XSA form}
+\label{fig:xsa}
+\end{figure}
 
@@
                    caption={Pseudocode algorithm for inserting beta nodes},
-                   label={lst:add_betas}, float=ht]
+                   label={lst:add_betas}]
 def addBetaNodesToFunc(cfg):
@@
                    caption={Pseudocode algorithm for propagating range information},
-                   label={lst:propagate_ranges}, float=ht]
+                   label={lst:propagate_ranges}]
 def propagateRanges(graph):
@@
 There are three benchmark suites which are generally used to measure JavaScript
-JIT compilers: the v8bench suite developed by Google\cite{v8bench}, the SunSpider benchmark
-suite developed by Apple\cite{sunspider}, and the Kraken benchmark suite developed by Mozilla\cite{kraken}
+JIT compilers: the v8bench suite developed by Google, the SunSpider benchmark
+suite developed by Apple, and the Kraken benchmark suite developed by Mozilla.
 
@@
 analysis pass, and measured the performance differences. The results are
-summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and
-\ref{fig:kraken}. These figures were gathered using the SunSpider test harness,
-which gives both a mean and a 95\% confidence interval\cite{sunspider}. Each
-test was run 30 times.
+summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and \ref{fig:kraken}.
 

### 2012-05-02 project-writings/paper.tex
explain why we don't DCE

--- 2012-05-02 project-writings/paper.tex
@@
 
-\section{Surprises and Lessons Learned}
-%% XXX: TODO
-%% Were any of your results surprising our counter-intuitive? If so, what did you learn from those surprises? What did you learn from this experience in general?
-Boy, JS is a pain in the ass.
-This is a lesson learned, not a surprise.
-
 \section{Future Work and Conclusions}
@@
 
-A number of further optimizations are possible using our range
-analysis framework. If the range associated with an instruction
-becomes empty, then the block has been proven dead and can be
-eliminated. Control flow operations can be proven redundant as well,
-further eliminating computation and dead basic blocks. These graph
-manipulations are currently not available in the IonMonkey compiler,
-which is in active development but not yet complete.
+A number of Further optimizations are possible using our range analysis
+framework. If the range associated with an instruction becomes empty, then the
+block has been proven dead and can be eliminated. Control flow operations can
+be proven redundant as well, further eliminating computation and dead basic
+blocks. These graph manipulations are currently not available in the IonMonkey
+compiler, which is in active development but not yet complete. There are a
+variety of reasons for this current design restriction. IonMonkey must
+integrate very closely with the interpreter and the JS runtime, and so
+generated code must maintain a careful correspondence with the interpreter.
+This means that eliminating control flow is a non-trivial
+operation\cite{dvander}.
 

### 2012-05-02 project-writings/paper.tex
talk about strict v1 < v2 comparisons

@@
 compared to a constant, we compute a range and use it to create a new
-$\beta$ node. We want to propagate this range information to all uses
+$\beta$ node. As a further optimization, we furthermore insert trivial beta
+nodes when we encounter a
+strict comparison between two variables--the larger must be at least
+$-2^{31}+1$, and the smaller can be at most $2^{31}-2$.
+
+After inserting a $\beta$ node, we then want to propagate this range information to all uses
 of the variable where the branch is guaranteed to have been taken;

### 2012-05-02 project-writings/paper.tex
Fix some mistakes.

--- 2012-05-02 project-writings/paper.tex
@@
 In order to get a different view of how well our optimization was
-performing, we annotated IonMonkey to emit for each addition
+performing, we annotated IonMonkey to emit for each addition and subtraction
 instruction compiled whether or not range analysis proved that it
@@
 \includegraphics[width=0.98\textwidth]{counts_graph.png}
-\caption{Percentage of add overflow checks eliminated}
+\caption{Fraction of add/subtract overflow checks eliminated}
 \label{fig:count}
@@
 succeeds in eliminating a large fraction of overflow checks for
-integer additions. This translates into a minor overall performance
-win on the Kraken benchmark suite and a very minor overall performance
-loss on the Sunspider suite. Since we should only increase the
-quality of the generated code, any performance losses should be from
-increased compilation overhead (Sunspider tests in particular are
-quite short). The performance improvements are decent, but not
-stellar; we believe this is because Javascript code is generally not
-integer-arithmetic-heavy tight loops and because the overflow test
-branches are generally predicted correctly.
+integer additions and subtractions. This translates into a minor
+overall performance win on the Kraken benchmark suite and a very minor
+overall performance loss on the Sunspider suite. Since we should only
+increase the quality of the generated code, any performance losses
+should be from increased compilation overhead (Sunspider tests in
+particular are quite short). The performance improvements are decent,
+but not stellar; we believe this is because Javascript code is
+generally not integer-arithmetic-heavy tight loops and because the
+overflow test branches are generally predicted correctly.
 

### 2012-05-02 project-writings/paper.tex
munge the tables around so latex doesn't deem it appropriate to just have one tiny table occupying all of page 7.
yield -> not yield

@@
 test was run 30 times.
-
 \begin{figure}[H]
+\begin{center}
 \begin{tabular}{|l|l|l|l|l|}
@@
 \label{fig:v8bench}
+\end{center}
 \end{figure}
@@
 \begin{figure}[H]
+\begin{tabular}{|l|l|l|l|l|}
+\hline
+    \textbf{Test}     &\textbf{Comparison} & \textbf{From}        & \textbf{To}           &  \textbf{Details} \\
+\hline\hline
+    Total:                      & 1.015x as fast   & 2319.6ms $\pm$ 0.3\% &  2285.4ms $\pm$ 0.4\%  &   significant\\
+\hline\hline
+      ai:                       & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
+\hline\hspace{0.5em} astar:                  & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
+\hline\hline
+      audio:                    & -                &  842.8ms $\pm$ 0.4\% &   838.0ms $\pm$ 0.7\%  &\\
+\hline\hspace{0.5em} beat-detection:         & ??               &  393.0ms $\pm$ 1.0\% &   397.6ms $\pm$ 1.5\%  &   \textit{might} be 1.012x as slow\\
+\hline\hspace{0.5em} dft:                    & 1.030x as fast   &  162.6ms $\pm$ 0.6\% &   157.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{0.5em} fft:                    & 1.016x as fast   &  123.5ms $\pm$ 0.2\% &   121.6ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} oscillator:             & 1.017x as fast   &  163.7ms $\pm$ 0.3\% &   160.9ms $\pm$ 0.3\%  &   significant\\
+\hline\hline
+      imaging:                  & 1.020x as fast   &  569.1ms $\pm$ 0.9\% &   557.9ms $\pm$ 1.3\%  &   significant\\
+\hline\hspace{0.5em} gaussian-blur:          & -                &  197.2ms $\pm$ 1.7\% &   193.7ms $\pm$ 2.1\%  &\\
+\hline\hspace{0.5em} darkroom:               & -                &  248.2ms $\pm$ 0.3\% &   247.6ms $\pm$ 0.4\%  &\\
+\hline\hspace{0.5em} desaturate:             & 1.062x as fast   &  123.8ms $\pm$ 3.2\% &   116.7ms $\pm$ 5.1\%  &   significant\\
+\hline\hline
+      json:                     & 1.019x as fast   &  121.3ms $\pm$ 0.2\% &   119.0ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} parse-financial:        & 1.005x as fast   &   75.8ms $\pm$ 0.2\% &    75.5ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} stringify-tinderbox:    & 1.044x as fast   &   45.5ms $\pm$ 0.5\% &    43.5ms $\pm$ 0.8\%  &   significant\\
+\hline\hline
+      stanford:                 & -                &  665.0ms $\pm$ 0.5\% &   662.2ms $\pm$ 0.4\%  &\\
+\hline\hspace{0.5em} crypto-aes:             & -                &  160.7ms $\pm$ 1.5\% &   160.6ms $\pm$ 1.1\%  &\\
+\hline\hspace{0.5em} crypto-ccm:             & 1.008x as slow &  158.6ms $\pm$ 0.4\% &   159.8ms $\pm$ 0.5\%  &   significant\\
+\hline\hspace{0.5em} crypto-pbkdf2:          & 1.006x as fast   &  254.9ms $\pm$ 0.5\% &   253.3ms $\pm$ 0.4\%  &   significant\\
+\hline\hspace{0.5em} crypto-sha256-iterative:& 1.025x as fast   &   90.8ms $\pm$ 0.4\% &    88.5ms $\pm$ 0.5\%  &   significant \\
+\hline
+\end{tabular}
+\caption{Kraken benchmark results}
+\label{fig:kraken}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
 \begin{tabular}{|l|l|l|l|l|}
@@
 \label{fig:sunspider}
-\end{figure}
-
-\begin{figure}[H]
-\begin{tabular}{|l|l|l|l|l|}
-\hline
-    \textbf{Test}     &\textbf{Comparison} & \textbf{From}        & \textbf{To}           &  \textbf{Details} \\
-\hline\hline
-    Total:                      & 1.015x as fast   & 2319.6ms $\pm$ 0.3\% &  2285.4ms $\pm$ 0.4\%  &   significant\\
-\hline\hline
-      ai:                       & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
-\hline\hspace{0.5em} astar:                  & 1.118x as fast   &  121.3ms $\pm$ 0.8\% &   108.5ms $\pm$ 1.2\%  &   significant\\
-\hline\hline
-      audio:                    & -                &  842.8ms $\pm$ 0.4\% &   838.0ms $\pm$ 0.7\%  &\\
-\hline\hspace{0.5em} beat-detection:         & ??               &  393.0ms $\pm$ 1.0\% &   397.6ms $\pm$ 1.5\%  &   \textit{might} be 1.012x as slow\\
-\hline\hspace{0.5em} dft:                    & 1.030x as fast   &  162.6ms $\pm$ 0.6\% &   157.8ms $\pm$ 0.5\%  &   significant\\
-\hline\hspace{0.5em} fft:                    & 1.016x as fast   &  123.5ms $\pm$ 0.2\% &   121.6ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{0.5em} oscillator:             & 1.017x as fast   &  163.7ms $\pm$ 0.3\% &   160.9ms $\pm$ 0.3\%  &   significant\\
-\hline\hline
-      imaging:                  & 1.020x as fast   &  569.1ms $\pm$ 0.9\% &   557.9ms $\pm$ 1.3\%  &   significant\\
-\hline\hspace{0.5em} gaussian-blur:          & -                &  197.2ms $\pm$ 1.7\% &   193.7ms $\pm$ 2.1\%  &\\
-\hline\hspace{0.5em} darkroom:               & -                &  248.2ms $\pm$ 0.3\% &   247.6ms $\pm$ 0.4\%  &\\
-\hline\hspace{0.5em} desaturate:             & 1.062x as fast   &  123.8ms $\pm$ 3.2\% &   116.7ms $\pm$ 5.1\%  &   significant\\
-\hline\hline
-      json:                     & 1.019x as fast   &  121.3ms $\pm$ 0.2\% &   119.0ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{0.5em} parse-financial:        & 1.005x as fast   &   75.8ms $\pm$ 0.2\% &    75.5ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{0.5em} stringify-tinderbox:    & 1.044x as fast   &   45.5ms $\pm$ 0.5\% &    43.5ms $\pm$ 0.8\%  &   significant\\
-\hline\hline
-      stanford:                 & -                &  665.0ms $\pm$ 0.5\% &   662.2ms $\pm$ 0.4\%  &\\
-\hline\hspace{0.5em} crypto-aes:             & -                &  160.7ms $\pm$ 1.5\% &   160.6ms $\pm$ 1.1\%  &\\
-\hline\hspace{0.5em} crypto-ccm:             & 1.008x as slow &  158.6ms $\pm$ 0.4\% &   159.8ms $\pm$ 0.5\%  &   significant\\
-\hline\hspace{0.5em} crypto-pbkdf2:          & 1.006x as fast   &  254.9ms $\pm$ 0.5\% &   253.3ms $\pm$ 0.4\%  &   significant\\
-\hline\hspace{0.5em} crypto-sha256-iterative:& 1.025x as fast   &   90.8ms $\pm$ 0.4\% &    88.5ms $\pm$ 0.5\%  &   significant \\
-\hline
-\end{tabular}
-\caption{Kraken benchmark results}
-\label{fig:kraken}
+\end{center}
 \end{figure}
@@
 successful in eliminating overflow checks from programs, but our
-timing results indicate that this does yield that substantial of a
+timing results indicate that this does not yield that substantial of a
 performance improvement. We believe that this is because the overhead

### 2012-05-02 project-writings/paper.tex
very minor nits

@@
 some measurable performance increase. In extremly short-running tests such
-Sunspider (see Figure \ref{fig:sunspider}), the additional compile-time
+SunSpider (see Figure \ref{fig:sunspider}), the additional compile-time
 overhead that our pass requires yields slightly slower performance.
@@
 could not overflow, and ran this against all of the tests in Kraken,
-Sunspider, and v8bench. The data from these tests are presented in Figure
+SunSpider, and v8bench. The data from these tests are presented in Figure
 \ref{fig:count} (tests that were found to have very few total add
@@
 performance improvement. We believe that this is because the overhead
-of an essentially never taken branch on modern hardware are quite
+of an essentially never taken branch on modern hardware is quite
 low. While on a microbenchmark we constructed overflow check
@@
 overall performance win on the Kraken benchmark suite and a very minor
-overall performance loss on the Sunspider suite. Since we should only
+overall performance loss on the SunSpider suite. Since we should only
 increase the quality of the generated code, any performance losses
-should be from increased compilation overhead (Sunspider tests in
+should be from increased compilation overhead (SunSpider tests in
 particular are quite short). The performance improvements are decent,
@@
 
-A number of Further optimizations are possible using our range analysis
+A number of further optimizations are possible using our range analysis
 framework. If the range associated with an instruction becomes empty, then the

### 2012-05-02 project-writings/paper.tex
fix ordering
typo

@@
 optimization by the commonly available benchmark suites.
-%% TODO : elim stats too
 
@@
 
-We ran each benchmark suite on IonMonkey, both with and without our range
-analysis pass, and measured the performance differences. The results are
-summarized in Figures \ref{fig:v8bench}, \ref{fig:sunspider}, and
-\ref{fig:kraken}. These figures were gathered using the SunSpider test harness,
-which gives both a mean and a 95\% confidence interval\cite{sunspider}. Each
-test was run 30 times.
+We ran each benchmark suite on IonMonkey, both with and without our
+range analysis pass, and measured the performance differences. The
+results are summarized in Figures \ref{fig:v8bench}, \ref{fig:kraken},
+and \ref{fig:sunspider}. These figures were gathered using the
+SunSpider test harness, which gives both a mean and a 95\% confidence
+interval\cite{sunspider}. Each test was run 30 times.
 \begin{figure}[H]
@@
 To summarize our evaluation, we discovered the checks we eliminate do result in
-some measurable performance increase. In extremly short-running tests such
+some measurable performance increase. In extremly short-running tests such as
 SunSpider (see Figure \ref{fig:sunspider}), the additional compile-time

### 2012-05-02 project-writings/paper.tex
Some fixes.

@@
 \item A different algorithm for inserting and removing $\beta$ nodes
-  that used by Gough and Klaeren that takes advantage of the data
+  than  used by Gough and Klaeren that takes advantage of the data
   structures of a modern SSA compiler.
@@
 An important source of range information that requires care to take
-advantage of is conditional branches. Consider the code in Figure
+advantage of is conditional branches. Consider the code in Listing
 \ref{fig:branch}. All three arithmetic operations in this code can be
@@
                    label={fig:branch}, float=ht]
-var y;
 if (x < 0) {
@@
 \begin{lstlisting}[language=Javascript,
-                   caption={Code from Figure \ref{fig:branch}, in XSA form},
+                   caption={Code from Listing \ref{fig:branch}, in XSA form},
                    label={fig:xsa}, float=ht]
@@
 do this, we take advantage of IonMonkey's use chains to efficiently
-iterate over all uses of an instruction.
+iterate over all uses of an instruction and modify them in place.
 
@@
 Recomputation of the possible range of an instruction is fairly simple
-and proceeds by the rules given in Figure \ref{fig:range_anal}. The
-key insight we had in these equations is that while the ranges of the
+and proceeds by the rules given in Figure \ref{fig:range_anal} (many
+operations that we support or omitted from the figure). The key
+insight we had in these equations is that while the ranges of the
 inputs to an operation may be infinite (that is, they may have
@@
 We do not perform general array bounds check elimination. To do this
-fully would require symbolic range analysis. Since IonMonkey already
-does hoisting of bounds checks out of loops, we determined that the
-effort would probably not be worth the gain.
+would in many cases require symbolic range analysis. Since IonMonkey
+already does hoisting of bounds checks out of loops, we determined
+that the effort would probably not be worth the gain.
 

### 2012-05-02 project-writings/paper.tex
More fixes

@@
 advantage of is conditional branches. Consider the code in Listing
-\ref{fig:branch}. All three arithmetic operations in this code can be
+\ref{lst:branch}. All three arithmetic operations in this code can be
 proven to not overflow, but to do this it is not sufficient to simply
@@
 branches with range information.
+Listing \ref{lst:xsa} shows the code from Listing \ref{lst:branch}
+lifted into XSA form.
 
@@
                    caption={Code that requires information about branches to properly analyze},
-                   label={fig:branch}, float=ht]
+                   label={lst:branch}, float=ht]
 if (x < 0) {
@@
 \begin{lstlisting}[language=Javascript,
-                   caption={Code from Listing \ref{fig:branch}, in XSA form},
-                   label={fig:xsa}, float=ht]
+                   caption={Code from Listing \ref{lst:branch}, in XSA form},
+                   label={lst:xsa}, float=ht]
 if (x < 0) {
@@
 
-At the start of range analysis, all ranges are initialized to
-$[-\infty, \infty]$.
-
 Our algorithm for range analsis is based on the algorithm due to Gough. It
@@
 
-We compute the range of each definition unconditionally once, and add all phi
+Pseudocode for range analysis is shown in Listing \ref{lst:propagate_ranges}.
+
+At the start of range analysis, all ranges are initialized to
+$[-\infty, \infty]$.
+We then
+compute the range of each definition unconditionally once, and add all phi
 nodes and beta nodes to a worklist--the only values which can change are those

### 2012-05-02 project-writings/paper.tex
fix wording about range analysis correctness
minor nits, done

@@
 
-\title{\bf Range Analysis for the IonMonkey JavaScript engine}
+\title{\bf Range Analysis for the IonMonkey JavaScript Compiler}
 
@@
 
-\subsection{Conditional Branches and $\beta$ nobes}
+\subsection{Conditional Branches and $\beta$ Nodes}
 
@@
 compared to a constant, we compute a range and use it to create a new
-$\beta$ node. As a further optimization, we furthermore insert trivial beta
+$\beta$ node. As an additional optimization, we furthermore insert trivial beta
 nodes when we encounter a
@@
 traversal is simpler to implement in IonMonkey, and in general definitions are
-still visited before all their uses. There is no issue of correctness since we
-iterate until no range information is changing.
+still visited before all their uses. There is no issue of correctness since at
+all points in the algorithm, the range information holds.
 

### 2012-05-03 project-writings/poster.tex
Start a poster.

\pagenumbering{gobble}
\documentclass{slides}
\usepackage{times}

\usepackage{amsmath,amsthm,amssymb,color}
%\usepackage{multicol}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{courier}
\usepackage{listings}
\usepackage{color}
%\usepackage[all]{hypcap}

\newcommand{\Squash}{\mathsf{Squash}}
\newcommand{\Clamp}{\mathsf{Clamp}}
\newcommand{\Range}{\mathsf{Range}}

\let\x\cdot

\begin{document}

\title{\bf Range Analysis for the IonMonkey JavaScript Compiler}
\author{Ryan Pearl and Michael Sullivan}

\maketitle

\begin{slide}
\begin{center}
{\Huge \bf{IonMonkey}}
\end{center}

\begin{itemize}
\item IonMonkey is Mozilla's next generation Javascript engine
\item Optimizing JIT compiler
\item Modern SSA-based compiler with internals inspired by LLVM
\item Performs optimistic type specialization and bails out when type
  assumptions fail
\item Designed to be easy to extend
\end{itemize}
\end{slide}

\begin{slide}
\begin{center}
{\Huge \bf{JavaScript arithmetic}}
\end{center}
\begin{itemize}
\item JavaScript numbers defined to be double precision floating point
\item For efficiency, implementations store and operate on them as
  integers whenever possible
\item IonMonkey will generates type specialized code for integer values
\item If arithmetic overflows, need to bail out of integer code and
  promote to doubles
\item Range analysis could eliminate these overflow checks
\end{itemize}
\end{slide}

\end{document}

### 2012-05-03 project-writings/poster.tex
more poster

@@
 \usepackage{graphicx}
+\usepackage{fullpage}
 \usepackage{hyperref}
@@
 \let\x\cdot
+\lstset{
+  language=Python,
+  showstringspaces=false,
+  tabsize=4,
+  basicstyle=\ttfamily,
+  keywordstyle=\bfseries\ttfamily,
+  captionpos=b
+}
+\lstdefinelanguage{JavaScript}{
+  keywords={if,else,var},
+  keywordstyle=\bfseries\ttfamily,
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/}
+}
 
@@
 \begin{center}
-{\Huge \bf{JavaScript arithmetic}}
+{\Huge \bf{JavaScript Arithmetic}}
 \end{center}
@@
 
+\begin{slide}
+\begin{center}
+{\Huge \bf{Range Analysis}}
+\end{center}
+\begin{itemize}
+\item Associate each integer temporary with a range: a contiguous subset of $\{-\infty\} \cup [-2^{31}, 2^{31}-1] \cup \{\infty\}$
+\item Can do arithmetic on ranges:
+\end{itemize}
+\begin{eqnarray*}
+\Clamp(n) &=& \begin{cases}
+-\infty &\text{if } n < -2^{31} \\
+\infty &\text{if } n > -2^{31}-1 \\
+n&\text{otherwise } \\
+\end{cases}\\
+\Clamp([l, h]) &=& [\Clamp(l), \Clamp(h)] \\
+{[x_l, x_h] \cup [y_l, y_h]} &=& [\min(x_l, y_l), \max(x_h, y_h)] \\
+{[x_l, x_h] \cap [y_l, y_h]} &=& [\max(x_l, y_l), \min(x_h, y_h)] \\
+{[x_l, x_h] + [y_l, y_h]} &=& \Clamp ([x_l + y_l, x_h + y_h]) \\
+{[x_l, x_h] - [y_l, y_h]} &=& \Clamp ([x_l - y_h, x_h - y_l])
+\end{eqnarray*}
+\end{slide}
+
+\begin{slide}
+\begin{center}
+{\Huge \bf{XSA Form}}
+\end{center}
+\begin{itemize}
+\item Want a way to track information learned from conditionals without needing to consider more than one range per temp
+\item Introduce $\beta$ nodes to associate range information from conditionals with temporaries
+\item Some example code with $\beta$ nodes; all of the ranges checks can be eliminated
+\end{itemize}
+{\tiny
+\begin{verbatim}
+if (x < 0) {
+    x1 = Beta(x, [INT_MIN, -1]);
+    y1 = x1 + 2000000000;
+} else {
+    x2 = Beta(x, [0, INT_MAX]);
+    if (x2 < 1000000000) {
+        x3 = Beta(x2, [INT_MIN, 999999999]);
+        y2 = x3*2;
+    } else {
+        x4 = Beta(x2, [1000000000, INT_MAX]);
+        y3 = x4 - 3000000000;
+    }
+    y4 = Phi(y2, y3);
+}
+y = Phi(y1, y4);
+\end{verbatim}
+}
+\end{slide}
+
 \end{document}

### 2012-05-03 project-writings/poster.tex
Finish up the slides

@@
 \item Associate each integer temporary with a range: a contiguous subset of $\{-\infty\} \cup [-2^{31}, 2^{31}-1] \cup \{\infty\}$
+\item Infinite ranges indicate the operation can overflow; arithmetic operations with a finite range don't need overflow checks!
 \item Can do arithmetic on ranges:
@@
 
+\begin{slide}
+\begin{center}
+{\Huge \bf{Range Analysis Algorithm}}
+\end{center}
+\begin{itemize}
+\item Range analysis is simple and worklist based. Range recomputation is done with range arithmetic. $\beta$ intersects a range, $\phi$ unions.
+\item Key insight: can always squash inputs to a finite range, since they would have bailed out if they overflowed.
+\end{itemize}
+{\tiny
+\begin{verbatim}
+def propagateRanges(graph):
+    worklist = []
+    for block in reversePostOrder(graph):
+        for insn in block:
+            insn.recomputeRange()
+            if !insn.isPhi() and !insn.isBeta():
+                continue
+            worklist.appendAll(insn.uses())
+
+    while len(worklist) > 0:
+        insn = worklist.pop()
+        #recomputeRange() returns true if the range changed
+        if (insn.recomputeRange()):
+            worklist.appendAll(insn.uses())
+\end{verbatim}
+}
+\end{slide}
+
+\begin{slide}
+\begin{center}
+{\Huge \bf{Results}}
+\end{center}
+\begin{itemize}
+\item Instrumented IonMonkey to output percentage of adds and
+  subtracts eliminated on a number of standard benchmark tests
+\item Encouraging results
+\end{itemize}
+\centering
+\includegraphics[width=0.98\textwidth]{counts_graph.png}
+\end{slide}
+
+\begin{slide}
+\begin{center}
+{\Huge \bf{Results }}
+\end{center}
+\begin{itemize}
+\item Tested with three benchmark suites: v8bench, Sunspider, and Kraken
+\item No statistically significant effect on v8bench
+\item Tiny performance regression on Sunspider (0.1\%)
+\item 1.5\% performance improvement on Kraken
+\item Real overall wins, but not much.
+\item Branch prediction is good these days, overflow checks wind up as
+  noise in real programs
+\item Microbenchmarks can see 30\% improvement
+\end{itemize}
+\end{slide}
+
 \end{document}
